---
title: "Grade 5 JMTES Previous/Current Merge"
author: "CSIC Evaluation Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

This R Markdown file creates the Grade 5 JMTES merged dataset by matching previous-year and current-year records on `Student ID`.
It uses a full outer join so students found in only one source sheet are retained.

For Grade 5, the previous-year data are from the `24-25 4th grade` sheet and the current-year data are from the `25-26 5th grade` sheet in the JMTES ATLAS workbook.

# Load packages

```{r packages}
library(readxl)
library(dplyr)
library(stringr)
library(writexl)
```

# Paths and helper functions

```{r helpers}
grade_dir <- "."
root_dir <- ".."
source_file <- file.path(root_dir, "JMTES ATLAS Report_Updated 5.22.26 (PW_ ATLAS).xlsx")
out_csv <- file.path(grade_dir, "JMTES Grade 5 Previous Current Merged.csv")
out_xlsx <- file.path(grade_dir, "JMTES Grade 5 Previous Current Merged.xlsx")

clean_id <- function(x) {
  id <- str_trim(as.character(x))
  id[id == ""] <- NA_character_
  sci_or_decimal <- !is.na(id) & (str_detect(str_to_upper(id), "E") | str_detect(id, "\\.0$"))
  id[sci_or_decimal] <- as.character(as.integer(as.numeric(id[sci_or_decimal])))
  str_replace_all(id, "\\D", "")
}

clean_header <- function(x) {
  x %>%
    as.character() %>%
    str_replace_all("\\n", " ") %>%
    str_squish() %>%
    str_replace("Student IDs", "Student ID")
}

make_headers <- function(raw, header_row_idx, prefix) {
  subject_row <- if (header_row_idx > 1) raw[header_row_idx - 1, ] else raw[0, ]
  header_row <- raw[header_row_idx, ]
  active_subject <- ""

  headers <- character(ncol(raw))
  for (idx in seq_len(ncol(raw))) {
    subject <- if (header_row_idx > 1) clean_header(subject_row[[idx]]) else ""
    field <- clean_header(header_row[[idx]])
    if (!is.na(subject) && subject != "") active_subject <- subject

    headers[[idx]] <- dplyr::case_when(
      is.na(field) | field == "" ~ "",
      field %in% c("Grade", "Teacher", "Student ID") ~ field,
      active_subject != "" ~ paste(prefix, active_subject, field),
      TRUE ~ paste(prefix, field)
    )
  }
  headers
}

extract_dataset <- function(sheet_name, prefix) {
  raw <- read_excel(source_file, sheet = sheet_name, col_names = FALSE, col_types = "text")
  header_row_idx <- which(apply(raw, 1, function(row) any(clean_header(row) == "Student ID")))[1]
  if (is.na(header_row_idx)) stop("No Student ID header row found in ", sheet_name)

  headers <- make_headers(raw, header_row_idx, prefix)
  id_idx <- which(headers == "Student ID")
  if (length(id_idx) != 1) stop("Could not identify one Student ID column in ", sheet_name)

  data <- raw[(header_row_idx + 1):nrow(raw), , drop = FALSE]
  names(data) <- make.unique(ifelse(headers == "", paste0("blank_", seq_along(headers)), headers))

  out <- data %>%
    mutate(`Student ID` = clean_id(.data[[names(data)[id_idx]]])) %>%
    filter(!is.na(`Student ID`), `Student ID` != "") %>%
    select(-starts_with("blank_")) %>%
    rename_with(~ paste(prefix, .x), any_of(c("Grade", "Teacher")))

  if (anyDuplicated(out$`Student ID`)) {
    stop("Duplicate Student ID found in ", sheet_name)
  }

  out
}
```

# Merge Grade 5 datasets

```{r merge-grade-5}
previous <- extract_dataset("24-25 4th grade", "Previous Year")
current <- extract_dataset("25-26 5th grade", "Current Year")

merged <- full_join(previous, current, by = "Student ID") %>%
  mutate(
    `Merge Status` = case_when(
      !is.na(`Previous Year Grade`) & !is.na(`Current Year Grade`) ~ "Both years",
      !is.na(`Previous Year Grade`) ~ "Previous year only",
      TRUE ~ "Current year only"
    ),
    .before = 2
  )

write.csv(merged, out_csv, row.names = FALSE, na = "")
write_xlsx(list("Merged Grade 5" = merged), out_xlsx)

merged %>% count(`Merge Status`)
```

# Outputs

```{r output-paths}
out_csv
out_xlsx
```
