---
title: "At Risk Report Absence Percentage Calculation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Purpose

This notebook calculates absence percentages for the JMTES and JES At Risk Report files using **170 school days** as the full-year denominator.

Running this notebook creates two Excel workbooks in the project folder:

- `JMTES At Risk Report with Absence Percentages.xlsx`
- `JES At Risk Report with Absence Percentages.xlsx`

The Excel workbooks are generated outputs and are intentionally not committed to Git because `.xlsx` files are binary files that cannot be displayed in pull-request diffs. To make the calculated results reviewable in GitHub, this notebook also writes CSV copies of the updated `Page1` sheets, which are committed with this change.

Because the **JMTES** report was pulled on **May 7, 2026**, before the May 22 end of the 170-day school year, its percentages are adjusted to a 170-day school-year rate using the 159 school days completed by the report date. In other words, JMTES absence counts are first annualized to the 170-day school year, then divided by 170. This is equivalent to dividing the May 7 absence counts by 159. The **JES** report was pulled after the school year ended, so it uses the full 170 days directly.

The added fields are:

- `Full School Year Days` = `170`
- `Report School Days` = school days represented in the report (`159` for JMTES; `170` for JES)
- `Excused Absences Adjusted to 170-Day Year` = `Excused / Report School Days * 170`
- `Unexcused Absences Adjusted to 170-Day Year` = `Unexcused / Report School Days * 170`
- `Excused Absence %` = `Excused Absences Adjusted to 170-Day Year / 170 * 100`
- `Unexcused Absence %` = `Unexcused Absences Adjusted to 170-Day Year / 170 * 100`
- `Total Absences` = `Excused + Unexcused`
- `Total Absences Adjusted to 170-Day Year` = `Total Absences / Report School Days * 170`
- `Total Absence %` = `Total Absences Adjusted to 170-Day Year / 170 * 100`

```{r libraries}
# Install these packages first if needed:
# install.packages(c("readxl", "writexl", "dplyr"))

library(readxl)
library(writexl)
library(dplyr)
```

```{r constants}
full_school_year_days <- 170
jmtes_report_school_days <- 159
jes_report_school_days <- 170

jmtes_input <- "JMTES At Risk Report_Updated 5.7.26 (PW_ GROW2026).xlsx"
jes_input <- "JES At Risk Report 25-26_Updated 5.28.26 (PW_ Titans2026) .xlsx"

jmtes_output <- "JMTES At Risk Report with Absence Percentages.xlsx"
jes_output <- "JES At Risk Report with Absence Percentages.xlsx"

jmtes_csv_output <- "JMTES At Risk Report with Absence Percentages.csv"
jes_csv_output <- "JES At Risk Report with Absence Percentages.csv"
```

```{r helper-functions}
add_absence_percentages <- function(input_file,
                                    output_file,
                                    report_school_days,
                                    full_school_year_days = 170) {
  at_risk <- read_excel(input_file, sheet = "Page1", skip = 1) %>%
    mutate(
      Excused = as.numeric(Excused),
      Unexcused = as.numeric(Unexcused),
      `Full School Year Days` = full_school_year_days,
      `Report School Days` = report_school_days,
      `Excused Absences Adjusted to 170-Day Year` = round(Excused / report_school_days * full_school_year_days, 2),
      `Unexcused Absences Adjusted to 170-Day Year` = round(Unexcused / report_school_days * full_school_year_days, 2),
      `Excused Absence %` = round(`Excused Absences Adjusted to 170-Day Year` / full_school_year_days * 100, 2),
      `Unexcused Absence %` = round(`Unexcused Absences Adjusted to 170-Day Year` / full_school_year_days * 100, 2),
      `Total Absences` = Excused + Unexcused,
      `Total Absences Adjusted to 170-Day Year` = round(`Total Absences` / report_school_days * full_school_year_days, 2),
      `Total Absence %` = round(`Total Absences Adjusted to 170-Day Year` / full_school_year_days * 100, 2)
    )

  code_sheet <- read_excel(input_file, sheet = "Code", col_names = FALSE)

  write_xlsx(
    list(
      Page1 = at_risk,
      Code = code_sheet
    ),
    path = output_file
  )

  at_risk
}
```

```{r calculate-and-write-files}
jmtes_absence <- add_absence_percentages(
  jmtes_input,
  jmtes_output,
  report_school_days = jmtes_report_school_days,
  full_school_year_days = full_school_year_days
)

jes_absence <- add_absence_percentages(
  jes_input,
  jes_output,
  report_school_days = jes_report_school_days,
  full_school_year_days = full_school_year_days
)

write.csv(jmtes_absence, jmtes_csv_output, row.names = FALSE)
write.csv(jes_absence, jes_csv_output, row.names = FALSE)
```

```{r confirm-output-files}
stopifnot(file.exists(jmtes_output))
stopifnot(file.exists(jes_output))
stopifnot(file.exists(jmtes_csv_output))
stopifnot(file.exists(jes_csv_output))

message("Created Excel files:")
message("- ", jmtes_output)
message("- ", jes_output)
message("Created reviewable CSV copies:")
message("- ", jmtes_csv_output)
message("- ", jes_csv_output)
```

```{r preview-results}
head(jmtes_absence)
head(jes_absence)
```
