---
title: "At Risk Report Absence Percentage Calculation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Purpose

This notebook calculates absence percentages for the JMTES and JES At Risk Report files using **170 school days** as the denominator.

Running this notebook creates two Excel workbooks in the project folder:

- `JMTES At Risk Report with Absence Percentages.xlsx`
- `JES At Risk Report with Absence Percentages.xlsx`

The Excel workbooks are generated outputs and are intentionally not committed to Git because `.xlsx` files are binary files that cannot be displayed in pull-request diffs. To make the calculated results reviewable in GitHub, this notebook also writes CSV copies of the updated `Page1` sheets, which are committed with this change.

The added fields are:

- `Excused Absence %` = `Excused / 170 * 100`
- `Unexcused Absence %` = `Unexcused / 170 * 100`
- `Total Absences` = `Excused + Unexcused`
- `Total Absence %` = `(Excused + Unexcused) / 170 * 100`

```{r libraries}
# Install these packages first if needed:
# install.packages(c("readxl", "writexl", "dplyr"))

library(readxl)
library(writexl)
library(dplyr)
```

```{r constants}
school_days <- 170

jmtes_input <- "JMTES At Risk Report_Updated 5.7.26 (PW_ GROW2026).xlsx"
jes_input <- "JES At Risk Report 25-26_Updated 5.28.26 (PW_ Titans2026) .xlsx"

jmtes_output <- "JMTES At Risk Report with Absence Percentages.xlsx"
jes_output <- "JES At Risk Report with Absence Percentages.xlsx"

jmtes_csv_output <- "JMTES At Risk Report with Absence Percentages.csv"
jes_csv_output <- "JES At Risk Report with Absence Percentages.csv"
```

```{r helper-functions}
add_absence_percentages <- function(input_file, output_file, school_days = 170) {
  at_risk <- read_excel(input_file, sheet = "Page1", skip = 1) %>%
    mutate(
      Excused = as.numeric(Excused),
      Unexcused = as.numeric(Unexcused),
      `Excused Absence %` = round(Excused / school_days * 100, 2),
      `Unexcused Absence %` = round(Unexcused / school_days * 100, 2),
      `Total Absences` = Excused + Unexcused,
      `Total Absence %` = round(`Total Absences` / school_days * 100, 2)
    )

  code_sheet <- read_excel(input_file, sheet = "Code", col_names = FALSE)

  write_xlsx(
    list(
      Page1 = at_risk,
      Code = code_sheet
    ),
    path = output_file
  )

  at_risk
}
```

```{r calculate-and-write-files}
jmtes_absence <- add_absence_percentages(jmtes_input, jmtes_output, school_days)
jes_absence <- add_absence_percentages(jes_input, jes_output, school_days)

write.csv(jmtes_absence, jmtes_csv_output, row.names = FALSE)
write.csv(jes_absence, jes_csv_output, row.names = FALSE)
```

```{r confirm-output-files}
stopifnot(file.exists(jmtes_output))
stopifnot(file.exists(jes_output))
stopifnot(file.exists(jmtes_csv_output))
stopifnot(file.exists(jes_csv_output))

message("Created Excel files:")
message("- ", jmtes_output)
message("- ", jes_output)
message("Created reviewable CSV copies:")
message("- ", jmtes_csv_output)
message("- ", jes_csv_output)
```

```{r preview-results}
head(jmtes_absence)
head(jes_absence)
```
