---
title: "Covariate-Adjusted Attendance Regression Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Purpose

This notebook implements the requested attendance methodology: a covariate-adjusted, regression-based comparison of JMTES, the GROW treatment school, and JES, the comparison school.

The outcomes are:

- unexcused absence rate, measured as the percentage of enrolled school days missed;
- excused absence rate, measured as the percentage of enrolled school days missed; and
- total absence rate, measured as the percentage of enrolled school days missed.

The coefficient on `JMTES` is the estimated GROW treatment-control difference. Negative values mean lower absence rates at JMTES than JES after adjusting for the included covariates.

## Why this regression is useful, and why overlap weighting was primary

This regression model is useful because it is transparent, easy to communicate, and matches a standard evaluation specification: regress the attendance outcome on a treatment indicator and student covariates.

However, the overlap-weighted propensity-score design remains the more defensible **primary** observational analysis because it explicitly emphasizes students who are comparable across JMTES and JES on observed covariates. Regression uses all students and relies more heavily on linear-model extrapolation if the two schools differ in covariate composition. In practice, both approaches point to the same conclusion, so this regression is included as a robustness check and as the direct implementation of the requested methodology.

```{r libraries}
# install.packages(c("dplyr", "sandwich", "lmtest")) # if needed
library(dplyr)
library(sandwich)
library(lmtest)
```

```{r load-data}
jmtes <- read.csv("JMTES At Risk Report with Absence Percentages.csv", check.names = FALSE)
jes <- read.csv("JES At Risk Report with Absence Percentages.csv", check.names = FALSE)

jmtes$school <- "JMTES"
jes$school <- "JES"

dat <- bind_rows(jmtes, jes)
dat$JMTES <- ifelse(dat$school == "JMTES", 1, 0)
```

```{r school-calendar}
school_closures <- as.Date(c(
  "2025-09-01",
  "2025-10-31",
  paste0("2025-11-", 24:28),
  paste0("2025-12-", 22:31),
  "2026-01-01",
  "2026-01-02",
  "2026-01-19",
  paste0("2026-01-", 26:30),
  "2026-02-02",
  paste0("2026-03-", 23:27),
  "2026-04-24"
))

count_school_days <- function(start_date, end_date) {
  start_date <- max(start_date, as.Date("2025-08-18"))
  end_date <- min(end_date, as.Date("2026-05-22"))

  if (start_date > end_date) {
    return(0)
  }

  days <- seq.Date(start_date, end_date, by = "day")
  sum(!(weekdays(days) %in% c("Saturday", "Sunday")) & !(days %in% school_closures))
}
```

```{r construct-regression-data}
to_num <- function(x) {
  x <- suppressWarnings(as.numeric(x))
  x[is.na(x)] <- 0
  x
}

excel_origin <- as.Date("1899-12-30")
dat$entry_date <- excel_origin + to_num(dat[["Entry Date - E/W"]])
dat$report_end <- ifelse(dat$school == "JMTES", "2026-05-07", "2026-05-22")
dat$report_end <- as.Date(dat$report_end)

dat$enrolled_days <- mapply(count_school_days, dat$entry_date, dat$report_end)

dat$unexcused <- to_num(dat[["Unexcused"]])
dat$excused <- to_num(dat[["Excused"]])
dat$total_absences <- dat$unexcused + dat$excused

dat$unexcused_absence_rate <- dat$unexcused / dat$enrolled_days * 100
dat$excused_absence_rate <- dat$excused / dat$enrolled_days * 100
dat$total_absence_rate <- dat$total_absences / dat$enrolled_days * 100

dat$Female <- ifelse(trimws(dat[["Gender"]]) == "Female", 1, 0)
dat$Black <- ifelse(trimws(dat[["Ethnic Name"]]) == "Black", 1, 0)

dat$SES <- case_when(
  trimws(dat[["Meal Status Code"]]) == "03" ~ 0,
  trimws(dat[["Meal Status Code"]]) %in% c("02", "06") ~ 1,
  TRUE ~ 2
)

dat$grade <- factor(dat[["Student Grade"]])
```

```{r regression-helper}
run_attendance_model <- function(outcome) {
  formula <- as.formula(
    paste0(outcome, " ~ JMTES + Female + Black + SES + grade")
  )

  model <- lm(formula, data = dat)
  robust_results <- coeftest(model, vcov = vcovHC(model, type = "HC1"))
  treatment_row <- robust_results["JMTES", ]

  data.frame(
    outcome = outcome,
    estimand = "JMTES - JES",
    coefficient_jmtes = treatment_row["Estimate"],
    robust_se = treatment_row["Std. Error"],
    ci_low = treatment_row["Estimate"] - 1.96 * treatment_row["Std. Error"],
    ci_high = treatment_row["Estimate"] + 1.96 * treatment_row["Std. Error"],
    p_value = treatment_row["Pr(>|t|)"],
    students = nobs(model)
  )
}
```

```{r fit-models}
attendance_regression_results <- bind_rows(
  run_attendance_model("unexcused_absence_rate"),
  run_attendance_model("excused_absence_rate"),
  run_attendance_model("total_absence_rate")
)

outcome_labels <- c(
  unexcused_absence_rate = "Unexcused absence rate (% enrolled days)",
  excused_absence_rate = "Excused absence rate (% enrolled days)",
  total_absence_rate = "Total absence rate (% enrolled days)"
)

attendance_regression_results$outcome <- outcome_labels[attendance_regression_results$outcome]

write.csv(
  dat,
  "grow_attendance_regression_student_data.csv",
  row.names = FALSE
)

write.csv(
  attendance_regression_results,
  "grow_attendance_regression_results.csv",
  row.names = FALSE
)
```

```{r results}
attendance_regression_results
```

## Results

The regression results support the same broad conclusion as the overlap-weighted analysis:

- JMTES has lower unexcused absence rates than JES after adjustment.
- JMTES has lower total absence rates than JES after adjustment.
- JMTES has slightly higher excused absence rates than JES after adjustment.

In the generated regression output, the JMTES coefficient is:

- **-4.44 percentage points** for unexcused absence rate;
- **+0.78 percentage points** for excused absence rate; and
- **-3.67 percentage points** for total absence rate.

These coefficients are interpreted as adjusted percentage-point differences in the percentage of enrolled school days missed.
