---
title: "Grade 4 JMTES Previous/Current Merge"
author: "CSIC Evaluation Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

This R Markdown file creates the Grade 4 JMTES merged dataset by matching previous-year and current-year records on `Student ID`.
It uses a full outer join so students found in only one source file are retained.

Expected source files: place exactly one previous-year JMTES `.xlsx` file and one current-year JMTES `.xlsx` file in this `Grade 4` folder. The filenames should include markers such as `Previous`/`Current`, `Prev`/`Curr`, or `24-25`/`25-26`.

# Load packages

```{r packages}
library(readxl)
library(dplyr)
library(stringr)
library(writexl)
```

# Paths and helper functions

```{r helpers}
grade_dir <- "."
out_csv <- file.path(grade_dir, "JMTES Grade 4 Previous Current Merged.csv")
out_xlsx <- file.path(grade_dir, "JMTES Grade 4 Previous Current Merged.xlsx")

clean_id <- function(x) {
  id <- str_trim(as.character(x))
  id[id == ""] <- NA_character_
  sci_or_decimal <- !is.na(id) & (str_detect(str_to_upper(id), "E") | str_detect(id, "\\.0$"))
  id[sci_or_decimal] <- as.character(as.integer(as.numeric(id[sci_or_decimal])))
  str_replace_all(id, "\\D", "")
}

clean_header <- function(x) {
  x %>%
    as.character() %>%
    str_replace_all("\\n", " ") %>%
    str_squish() %>%
    str_replace("Student IDs", "Student ID")
}

find_source_files <- function() {
  candidates <- list.files(
    grade_dir,
    pattern = "\\.xlsx$",
    full.names = TRUE,
    ignore.case = TRUE
  )
  candidates <- candidates[str_detect(basename(candidates), regex("jmtes", ignore_case = TRUE))]
  candidates <- candidates[!str_detect(basename(candidates), regex("merged", ignore_case = TRUE))]

  previous <- candidates[str_detect(basename(candidates), regex("previous|prev|24-25|2024", ignore_case = TRUE))]
  current <- candidates[str_detect(basename(candidates), regex("current|curr|25-26|2025", ignore_case = TRUE))]

  if (length(previous) != 1 || length(current) != 1 || previous == current) {
    stop(
      "Expected exactly one previous-year and one current-year JMTES .xlsx file in Grade 4. Found: ",
      paste(basename(candidates), collapse = ", ")
    )
  }

  list(previous = previous, current = current)
}

make_headers <- function(raw, header_row_idx, prefix) {
  subject_row <- if (header_row_idx > 1) raw[header_row_idx - 1, ] else raw[0, ]
  header_row <- raw[header_row_idx, ]
  active_subject <- ""

  headers <- character(ncol(raw))
  for (idx in seq_len(ncol(raw))) {
    subject <- if (header_row_idx > 1) clean_header(subject_row[[idx]]) else ""
    field <- clean_header(header_row[[idx]])
    if (!is.na(subject) && subject != "") active_subject <- subject

    headers[[idx]] <- dplyr::case_when(
      is.na(field) | field == "" ~ "",
      field %in% c("Grade", "Teacher", "Student ID") ~ field,
      active_subject != "" ~ paste(prefix, active_subject, field),
      TRUE ~ paste(prefix, field)
    )
  }
  headers
}

extract_dataset <- function(path, prefix) {
  sheets <- excel_sheets(path)

  for (sheet_name in sheets) {
    raw <- read_excel(path, sheet = sheet_name, col_names = FALSE, col_types = "text")
    header_row_idx <- which(apply(raw, 1, function(row) any(clean_header(row) == "Student ID")))[1]
    if (is.na(header_row_idx)) next

    headers <- make_headers(raw, header_row_idx, prefix)
    id_idx <- which(headers == "Student ID")
    if (length(id_idx) != 1) next

    data <- raw[(header_row_idx + 1):nrow(raw), , drop = FALSE]
    names(data) <- make.unique(ifelse(headers == "", paste0("blank_", seq_along(headers)), headers))

    out <- data %>%
      mutate(`Student ID` = clean_id(.data[[names(data)[id_idx]]])) %>%
      filter(!is.na(`Student ID`), `Student ID` != "") %>%
      select(-starts_with("blank_")) %>%
      rename_with(~ paste(prefix, .x), any_of(c("Grade", "Teacher")))

    if (anyDuplicated(out$`Student ID`)) {
      stop("Duplicate Student ID found in ", basename(path), " / ", sheet_name)
    }

    return(out)
  }

  stop("No worksheet with a Student ID column was found in ", basename(path))
}
```

# Merge Grade 4 datasets

```{r merge-grade-4}
files <- find_source_files()

previous <- extract_dataset(files$previous, "Previous Year")
current <- extract_dataset(files$current, "Current Year")

merged <- full_join(previous, current, by = "Student ID") %>%
  mutate(
    `Merge Status` = case_when(
      !is.na(`Previous Year Grade`) & !is.na(`Current Year Grade`) ~ "Both years",
      !is.na(`Previous Year Grade`) ~ "Previous year only",
      TRUE ~ "Current year only"
    ),
    .before = 2
  )

write.csv(merged, out_csv, row.names = FALSE, na = "")
write_xlsx(list("Merged Grade 4" = merged), out_xlsx)

merged %>% count(`Merge Status`)
```

# Outputs

```{r output-paths}
out_csv
out_xlsx
```
