---
title: "Grade 2 ATLAS Assessment Analysis: G.R.O.W Impact at JMTES"
author: "CSIC Evaluation Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 5.5)
```

# Purpose

This notebook mirrors the Kindergarten and Grade 1 ATLAS impact-analysis workflow for Grade 2.

- **Treatment school:** JMTES / Murrell Taylor Elementary, which received G.R.O.W.
- **Control school:** JES, which did not receive G.R.O.W.
- **Primary outcome:** ATLAS assessment scale scores.
- **Subjects:** ELA, Math.
- **Primary analysis:** ANCOVA comparing Summer 2026 scores between JMTES and JES while adjusting for the strongest common baseline score.
- **Sensitivity analyses:** DiD-style pre/post comparison, gain-score models, robust standard errors, Wilcoxon tests, winsorized gain-score models, calipered nearest-neighbor matching, inverse probability weighting, permutation inference, and bootstrap confidence intervals.

For Grade 2, Summer 2025 is available for both JMTES and JES and is the prior-year summative assessment before the 2025-26 outcome. Therefore, the primary ANCOVA uses Summer 2025 as the baseline.

# Load packages

```{r packages}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(broom)
library(knitr)
library(sandwich)
library(lmtest)
```

# File paths and helper functions

```{r paths-and-helpers}
jmtes_file <- "../JMTES ATLAS Report_Updated 5.22.26 (PW_ ATLAS).xlsx"
jes_file   <- "../JES ATLAS Report 25-26_Updated 5.28.26 (PW Titans2026).xlsx"

clean_score <- function(x) {
  score <- suppressWarnings(as.numeric(x))
  ifelse(score < 900, NA_real_, score)
}

subject_order <- c("ELA", "Math")
wave_order <- c("Summer 2025", "Winter 2025", "Summer 2026")

baseline_wave_for_subject <- function(subject) {
  ifelse(subject == "Science", "Winter 2025", "Summer 2025")
}
```

# Import and reshape ATLAS data

```{r import-data}
jmtes_raw <- read_excel(jmtes_file, sheet = "25-26 2nd grade", skip = 1, col_names = FALSE, col_types = "text")

jmtes_long <- jmtes_raw %>%
  transmute(grade = ...2, teacher = ...3, student_id = ...4,
    ela_sum_2025 = ...5, ela_winter_2025 = ...6, ela_sum_2026 = ...7,
    math_sum_2025 = ...8, math_winter_2025 = ...9, math_sum_2026 = ...10) %>%
  filter(!is.na(grade), grade != "", grade != "Grade") %>%
  mutate(school = "JMTES", treatment = 1L) %>%
  pivot_longer(cols = c(ela_sum_2025, ela_winter_2025, ela_sum_2026, math_sum_2025, math_winter_2025, math_sum_2026), names_to = "assessment", values_to = "score", values_drop_na = FALSE) %>%
  mutate(
    score = clean_score(score),
    subject = case_when(str_detect(assessment, "^ela") ~ "ELA", str_detect(assessment, "^math") ~ "Math", str_detect(assessment, "^science") ~ "Science", TRUE ~ NA_character_),
    wave = case_when(str_detect(assessment, "sum_2025") ~ "Summer 2025", str_detect(assessment, "fall_2025") ~ "Fall 2025", str_detect(assessment, "winter_2025") ~ "Winter 2025", str_detect(assessment, "sum_2026") ~ "Summer 2026", TRUE ~ NA_character_)
  ) %>%
  select(school, treatment, grade, teacher, student_id, subject, wave, score)

jes_raw <- read_excel(jes_file, sheet = "2nd", skip = 2, col_names = FALSE, col_types = "text")

jes_long <- jes_raw %>%
  transmute(grade = ...1, student_id = ...2,
    ela_sum_2025 = ...3, ela_winter_2025 = ...5, ela_sum_2026 = ...6,
    math_sum_2025 = ...7, math_winter_2025 = ...9, math_sum_2026 = ...10) %>%
  filter(!is.na(grade), grade != "", grade != "Grade") %>%
  mutate(school = "JES", treatment = 0L, teacher = NA_character_) %>%
  pivot_longer(cols = c(ela_sum_2025, ela_winter_2025, ela_sum_2026, math_sum_2025, math_winter_2025, math_sum_2026), names_to = "assessment", values_to = "score", values_drop_na = FALSE) %>%
  mutate(
    score = clean_score(score),
    subject = case_when(str_detect(assessment, "^ela") ~ "ELA", str_detect(assessment, "^math") ~ "Math", str_detect(assessment, "^science") ~ "Science", TRUE ~ NA_character_),
    wave = case_when(str_detect(assessment, "sum_2025") ~ "Summer 2025", str_detect(assessment, "fall_2025") ~ "Fall 2025", str_detect(assessment, "winter_2025") ~ "Winter 2025", str_detect(assessment, "sum_2026") ~ "Summer 2026", TRUE ~ NA_character_)
  ) %>%
  select(school, treatment, grade, teacher, student_id, subject, wave, score)

atlas_long <- bind_rows(jmtes_long, jes_long) %>%
  filter(!is.na(subject), !is.na(wave), !is.na(score)) %>%
  mutate(school = factor(school, levels = c("JES", "JMTES")), subject = factor(subject, levels = subject_order), wave = factor(wave, levels = wave_order))

wide_common <- atlas_long %>%
  select(school, treatment, student_id, subject, wave, score) %>%
  pivot_wider(names_from = wave, values_from = score) %>%
  mutate(
    baseline_wave = baseline_wave_for_subject(as.character(subject)),
    baseline = if_else(baseline_wave == "Winter 2025", `Winter 2025`, `Summer 2025`),
    post = `Summer 2026`,
    gain = post - baseline
  ) %>%
  filter(!is.na(baseline), !is.na(post))

kable(head(atlas_long, 10), caption = "First 10 standardized Grade 2 ATLAS observations")
```

# Data availability and mean scores

```{r descriptive}
data_availability <- atlas_long %>% count(school, subject, wave, name = "n_scores") %>% arrange(subject, school, wave)
mean_scores <- atlas_long %>%
  group_by(school, treatment, subject, wave) %>%
  summarise(n = n(), mean_score = mean(score), sd_score = sd(score), se_score = sd_score / sqrt(n), .groups = "drop") %>%
  arrange(subject, wave, school)

kable(data_availability, caption = "Available Grade 2 score counts by school, subject, and wave")
kable(mean_scores, digits = 2, caption = "Mean Grade 2 ATLAS scores by school, subject, and wave")
```

# Line graph by subject

```{r line-graphs, fig.height=6.5}
mean_line_plot <- ggplot(mean_scores, aes(x = wave, y = mean_score, color = school, group = school)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_score - 1.96 * se_score, ymax = mean_score + 1.96 * se_score), width = 0.08) +
  facet_wrap(~ subject, scales = "free_y") +
  labs(title = "Mean Grade 2 ATLAS Scores by School and Wave", subtitle = "JMTES received G.R.O.W; JES is the control school", x = "Assessment wave", y = "Mean ATLAS score", color = "School") +
  theme_minimal(base_size = 12) + theme(legend.position = "bottom")
mean_line_plot

dir.create("results", showWarnings = FALSE)
ggsave("results/grade2_atlas_mean_scores_line_plot.png", mean_line_plot, width = 9, height = 6.5, dpi = 300)
```

# Primary impact analysis: ANCOVA

```{r primary-ancova}
run_ancova <- function(data) { lm(post ~ treatment + baseline, data = data) }

ancova_models <- wide_common %>% group_by(subject) %>% group_split() %>% setNames(levels(droplevels(wide_common$subject))) %>% lapply(run_ancova)
ancova_results <- bind_rows(lapply(names(ancova_models), function(s) tidy(ancova_models[[s]], conf.int = TRUE) %>% mutate(subject = s, .before = 1)))
ancova_effects <- ancova_results %>% filter(term == "treatment") %>% transmute(subject, ancova_estimate = estimate, std_error = std.error, statistic, p_value = p.value, conf_low = conf.low, conf_high = conf.high)
kable(ancova_results, digits = 3, caption = "Primary Grade 2 ANCOVA results by subject")
kable(ancova_effects, digits = 3, caption = "Adjusted JMTES-vs-JES difference from Grade 2 ANCOVA model")
```

# Sensitivity and robustness checks

```{r sensitivity-models}
did_data <- wide_common %>% select(subject, school, treatment, baseline, post) %>% pivot_longer(cols = c(baseline, post), names_to = "period", values_to = "score") %>% mutate(post_period = if_else(period == "post", 1L, 0L), treatment_post = treatment * post_period)
did_models <- did_data %>% group_by(subject) %>% group_split() %>% setNames(levels(droplevels(did_data$subject))) %>% lapply(function(data) lm(score ~ treatment + post_period + treatment_post, data = data))
did_effects <- bind_rows(lapply(names(did_models), function(s) tidy(did_models[[s]], conf.int = TRUE) %>% mutate(subject = s, .before = 1))) %>% filter(term == "treatment_post")

gain_models <- wide_common %>% group_by(subject) %>% group_split() %>% setNames(levels(droplevels(wide_common$subject))) %>% lapply(function(data) lm(gain ~ treatment, data = data))
gain_results <- bind_rows(lapply(names(gain_models), function(s) tidy(gain_models[[s]], conf.int = TRUE) %>% mutate(subject = s, .before = 1)))
gain_summary <- wide_common %>% group_by(subject, school) %>% summarise(n = n(), mean_gain = mean(gain), sd_gain = sd(gain), .groups = "drop")

robust_did_results <- bind_rows(lapply(names(did_models), function(s) {
  robust_test <- coeftest(did_models[[s]], vcov. = vcovHC(did_models[[s]], type = "HC3"))
  tibble(subject = s, term = rownames(robust_test), estimate = robust_test[, 1], robust_se = robust_test[, 2], statistic = robust_test[, 3], p_value = robust_test[, 4])
}))

wilcoxon_results <- wide_common %>% group_by(subject) %>% summarise(n_jes = sum(school == "JES"), n_jmtes = sum(school == "JMTES"), wilcox_p_value = wilcox.test(gain ~ school, exact = FALSE)$p.value, .groups = "drop")

winsorize <- function(x, probs = c(0.05, 0.95)) { bounds <- quantile(x, probs = probs, na.rm = TRUE); pmin(pmax(x, bounds[[1]]), bounds[[2]]) }
wide_winsor <- wide_common %>% group_by(subject) %>% mutate(gain_winsor = winsorize(gain)) %>% ungroup()
winsor_models <- wide_winsor %>% group_by(subject) %>% group_split() %>% setNames(levels(droplevels(wide_winsor$subject))) %>% lapply(function(data) lm(gain_winsor ~ treatment, data = data))
winsor_results <- bind_rows(lapply(names(winsor_models), function(s) tidy(winsor_models[[s]], conf.int = TRUE) %>% mutate(subject = s, .before = 1)))

kable(did_effects, digits = 3, caption = "DiD-style sensitivity estimates")
kable(gain_summary, digits = 2, caption = "Gain summaries")
kable(gain_results, digits = 3, caption = "Gain-score model")
kable(robust_did_results, digits = 3, caption = "HC3 robust SE results")
kable(wilcoxon_results, digits = 3, caption = "Wilcoxon tests on gains")
kable(winsor_results, digits = 3, caption = "Winsorized gain-score model")
```

# Preferred matching method

This section mirrors Kindergarten and Grade 1: exact-by-subject, nearest-neighbor matching on the strongest available baseline score, with replacement and a 0.20 pooled-SD caliper. The baseline is Summer 2025 for ELA/Math and Winter 2025 for Grade 3 Science.

```{r matching}
pooled_sd <- function(x_treated, x_control) {
  n_t <- length(x_treated); n_c <- length(x_control)
  sqrt(((n_t - 1) * var(x_treated, na.rm = TRUE) + (n_c - 1) * var(x_control, na.rm = TRUE)) / (n_t + n_c - 2))
}

caliper_match_subject <- function(data, caliper_sd = 0.20) {
  treated <- data %>% filter(treatment == 1) %>% arrange(student_id)
  controls <- data %>% filter(treatment == 0) %>% arrange(student_id)
  caliper_points <- caliper_sd * pooled_sd(treated$baseline, controls$baseline)
  bind_rows(lapply(seq_len(nrow(treated)), function(i) {
    baseline_distance <- abs(controls$baseline - treated$baseline[i])
    match_index <- which.min(baseline_distance)
    accepted_match <- baseline_distance[match_index] <= caliper_points
    tibble(subject = as.character(treated$subject[i]), treated_id = treated$student_id[i], control_id = if_else(accepted_match, controls$student_id[match_index], NA_character_), treated_pre = treated$baseline[i], control_pre = if_else(accepted_match, controls$baseline[match_index], NA_real_), treated_post = treated$post[i], control_post = if_else(accepted_match, controls$post[match_index], NA_real_), caliper_points = caliper_points, baseline_distance = baseline_distance[match_index], matched = accepted_match, matched_effect = if_else(accepted_match, treated$post[i] - controls$post[match_index], NA_real_))
  }))
}

matched_pairs <- wide_common %>% group_by(subject) %>% group_split() %>% lapply(caliper_match_subject) %>% bind_rows()
matching_results <- matched_pairs %>% group_by(subject) %>% summarise(n_treated_total = n(), n_matched = sum(matched), n_unmatched = sum(!matched), caliper_points = first(caliper_points), mean_baseline_distance = mean(baseline_distance[matched], na.rm = TRUE), matched_effect = mean(matched_effect, na.rm = TRUE), sd_effect = sd(matched_effect, na.rm = TRUE), se_effect = sd_effect / sqrt(n_matched), statistic = matched_effect / se_effect, p_value = 2 * pt(-abs(statistic), df = n_matched - 1), conf_low = matched_effect + qt(0.025, df = n_matched - 1) * se_effect, conf_high = matched_effect + qt(0.975, df = n_matched - 1) * se_effect, .groups = "drop")
matching_balance <- matched_pairs %>% filter(matched) %>% group_by(subject) %>% summarise(treated_pre_mean = mean(treated_pre, na.rm = TRUE), matched_control_pre_mean = mean(control_pre, na.rm = TRUE), mean_baseline_difference_after_matching = treated_pre_mean - matched_control_pre_mean, mean_absolute_baseline_distance = mean(abs(treated_pre - control_pre), na.rm = TRUE), .groups = "drop")

kable(matching_results, digits = 3, caption = "Calipered nearest-neighbor matching results")
kable(matching_balance, digits = 3, caption = "Baseline balance after matching")
```

# IPW, permutation, and bootstrap checks

```{r ipw-permutation-bootstrap}
run_ipw <- function(data) {
  propensity_model <- glm(treatment ~ baseline, data = data, family = binomial())
  data_ipw <- data %>% mutate(propensity = predict(propensity_model, type = "response"), propensity = pmin(pmax(propensity, 0.01), 0.99), ipw = if_else(treatment == 1, 1 / propensity, 1 / (1 - propensity)))
  lm(post ~ treatment, data = data_ipw, weights = ipw)
}
ipw_models <- wide_common %>% group_by(subject) %>% group_split() %>% setNames(levels(droplevels(wide_common$subject))) %>% lapply(run_ipw)
ipw_results <- bind_rows(lapply(names(ipw_models), function(s) tidy(ipw_models[[s]], conf.int = TRUE) %>% mutate(subject = s, .before = 1)))

permute_ancova <- function(data, n_perm = 2000, seed = 2026) {
  set.seed(seed); observed <- coef(lm(post ~ treatment + baseline, data = data))["treatment"]
  permuted <- replicate(n_perm, { data_perm <- data %>% mutate(treatment = sample(treatment, size = n(), replace = FALSE)); coef(lm(post ~ treatment + baseline, data = data_perm))["treatment"] })
  tibble(observed_effect = unname(observed), n_permutations = n_perm, permutation_p_value = mean(abs(permuted) >= abs(observed), na.rm = TRUE), permutation_mean = mean(permuted, na.rm = TRUE), permutation_sd = sd(permuted, na.rm = TRUE))
}
permutation_results <- wide_common %>% group_by(subject) %>% group_split() %>% setNames(levels(droplevels(wide_common$subject))) %>% lapply(permute_ancova) %>% bind_rows(.id = "subject")

bootstrap_ancova <- function(data, n_boot = 2000, seed = 2026) {
  set.seed(seed); observed <- coef(lm(post ~ treatment + baseline, data = data))["treatment"]
  boot_estimates <- replicate(n_boot, { boot_data <- data[sample(seq_len(nrow(data)), size = nrow(data), replace = TRUE), ]; coef(lm(post ~ treatment + baseline, data = boot_data))["treatment"] })
  tibble(observed_effect = unname(observed), n_bootstrap = n_boot, bootstrap_se = sd(boot_estimates, na.rm = TRUE), conf_low = quantile(boot_estimates, 0.025, na.rm = TRUE), conf_high = quantile(boot_estimates, 0.975, na.rm = TRUE))
}
bootstrap_results <- wide_common %>% group_by(subject) %>% group_split() %>% setNames(levels(droplevels(wide_common$subject))) %>% lapply(bootstrap_ancova) %>% bind_rows(.id = "subject")

kable(ipw_results, digits = 3, caption = "IPW results")
kable(permutation_results, digits = 3, caption = "Permutation tests")
kable(bootstrap_results, digits = 3, caption = "Bootstrap confidence intervals")
```

# Methods considered but not used as primary analyses

The following common impact methods are not recommended as primary analyses with the current files: full causal DiD/event study, student fixed effects, school fixed effects with clustered standard errors, regression discontinuity, instrumental variables, and synthetic control. These approaches would require more pre-treatment waves, more schools, a valid cutoff, a valid instrument, or a larger control pool.

# Save analytic outputs

```{r save-outputs}
dir.create("results", showWarnings = FALSE)
write.csv(atlas_long, "results/grade2_atlas_long_cleaned.csv", row.names = FALSE)
write.csv(mean_scores, "results/grade2_atlas_mean_scores.csv", row.names = FALSE)
write.csv(ancova_effects, "results/grade2_atlas_ancova_effects.csv", row.names = FALSE)
write.csv(did_effects, "results/grade2_atlas_did_effects.csv", row.names = FALSE)
write.csv(gain_results, "results/grade2_atlas_gain_results.csv", row.names = FALSE)
write.csv(gain_summary, "results/grade2_atlas_gain_summary.csv", row.names = FALSE)
write.csv(robust_did_results, "results/grade2_atlas_robust_did_results.csv", row.names = FALSE)
write.csv(wilcoxon_results, "results/grade2_atlas_wilcoxon_results.csv", row.names = FALSE)
write.csv(winsor_results, "results/grade2_atlas_winsor_results.csv", row.names = FALSE)
write.csv(matched_pairs, "results/grade2_atlas_matching_pairs.csv", row.names = FALSE)
write.csv(matching_results, "results/grade2_atlas_matching_results.csv", row.names = FALSE)
write.csv(matching_balance, "results/grade2_atlas_matching_balance.csv", row.names = FALSE)
write.csv(ipw_results, "results/grade2_atlas_ipw_results.csv", row.names = FALSE)
write.csv(permutation_results, "results/grade2_atlas_permutation_results.csv", row.names = FALSE)
write.csv(bootstrap_results, "results/grade2_atlas_bootstrap_results.csv", row.names = FALSE)
```
