---
title: "Kindergarten ATLAS Assessment Analysis: G.R.O.W Impact at JMTES"
author: "CSIC Evaluation Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5.5
)
```

# Purpose

This notebook analyzes Kindergarten ATLAS assessment scores for the G.R.O.W evaluation.

- **Treatment school:** JMTES / Murrell Taylor Elementary, which received G.R.O.W.
- **Control school:** JES, which did not receive G.R.O.W.
- **Primary outcome:** ATLAS assessment scale scores.
- **Subjects:** ELA and Math.
- **Primary analysis:** ANCOVA, comparing Summer 2026 scores between JMTES and JES while adjusting for Winter 2025 baseline scores.
- **Sensitivity analyses:** DiD-style pre/post comparison, gain-score models, nearest-neighbor matching, inverse probability weighting, randomization/permutation inference, bootstrap confidence intervals, robust standard errors, Wilcoxon tests, and winsorized gain-score models.
- **Primary analytic question:** After adjusting for baseline Winter 2025 ATLAS scores, did Kindergarten students at the treatment school score higher than students at the control school in Summer 2026?

The strongest common pre/post comparison available in the two Kindergarten workbooks is:

- **Pre period:** Winter 2025
- **Post period:** Summer 2026

JES also includes Summer 2025 scores, but the JMTES Kindergarten workbook does not include Summer 2025. Therefore, the impact analysis focuses on the two waves available for both schools.

# Load packages

```{r packages}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(broom)
library(knitr)
library(sandwich)
library(lmtest)
```

# File paths and helper functions

```{r paths-and-helpers}
jmtes_file <- "JMTES Atlas Report - Kindergarten.xlsx"
jes_file   <- "JES Atlas Report - Kindergarten.xlsx"

clean_score <- function(x) {
  score <- suppressWarnings(as.numeric(x))
  # ATLAS scale scores in these files are in the 1000s. Values below 900,
  # such as 4/5 in JES Summer 2025, are screening/performance codes rather
  # than comparable ATLAS scale scores and are excluded from score analysis.
  ifelse(score < 900, NA_real_, score)
}

wave_order <- c("Summer 2025", "Winter 2025", "Summer 2026")
subject_order <- c("ELA", "Math")
```

# Import and reshape ATLAS data

The two school files use slightly different column names. This section standardizes the data into one long-format file with one row per student, school, subject, and wave.

```{r import-data}
# JMTES treatment school. Row 1 in this workbook is a title row, so data headers begin on row 2.
jmtes_raw <- read_excel(jmtes_file, skip = 1, col_types = "text")

jmtes_long <- jmtes_raw %>%
  rename(
    student_id = `Student ID`,
    grade = Grade,
    teacher = Teacher
  ) %>%
  mutate(
    school = "JMTES",
    treatment = 1L
  ) %>%
  pivot_longer(
    cols = matches("^(ELA|Math)"),
    names_to = "assessment",
    values_to = "score"
  ) %>%
  mutate(
    score = clean_score(score),
    subject = case_when(
      str_detect(assessment, "^ELA") ~ "ELA",
      str_detect(assessment, "^Math") ~ "Math",
      TRUE ~ NA_character_
    ),
    wave = case_when(
      str_detect(assessment, regex("Winter 2025", ignore_case = TRUE)) ~ "Winter 2025",
      str_detect(assessment, regex("Sum 2026|Summer 2026", ignore_case = TRUE)) ~ "Summer 2026",
      str_detect(assessment, regex("Sum 2025|Summer 2025", ignore_case = TRUE)) ~ "Summer 2025",
      TRUE ~ NA_character_
    )
  ) %>%
  select(school, treatment, grade, teacher, student_id, subject, wave, score)

# JES control school.
jes_raw <- read_excel(jes_file, col_types = "text")

jes_long <- jes_raw %>%
  rename(
    student_id = `ID Number`,
    grade = Grade
  ) %>%
  mutate(
    school = "JES",
    treatment = 0L,
    teacher = NA_character_
  ) %>%
  pivot_longer(
    cols = matches("^(ELA|Math)"),
    names_to = "assessment",
    values_to = "score"
  ) %>%
  mutate(
    score = clean_score(score),
    subject = case_when(
      str_detect(assessment, "^ELA") ~ "ELA",
      str_detect(assessment, "^Math") ~ "Math",
      TRUE ~ NA_character_
    ),
    wave = case_when(
      str_detect(assessment, regex("Winter 2025", ignore_case = TRUE)) ~ "Winter 2025",
      str_detect(assessment, regex("Sum 2026|Summer 2026", ignore_case = TRUE)) ~ "Summer 2026",
      str_detect(assessment, regex("Sum 2025|Summer 2025", ignore_case = TRUE)) ~ "Summer 2025",
      TRUE ~ NA_character_
    )
  ) %>%
  select(school, treatment, grade, teacher, student_id, subject, wave, score)

atlas_long <- bind_rows(jmtes_long, jes_long) %>%
  filter(!is.na(subject), !is.na(wave), !is.na(score)) %>%
  mutate(
    school = factor(school, levels = c("JES", "JMTES")),
    treatment = as.integer(treatment),
    subject = factor(subject, levels = subject_order),
    wave = factor(wave, levels = wave_order),
    post = if_else(wave == "Summer 2026", 1L, 0L)
  )

atlas_common <- atlas_long %>%
  filter(wave %in% c("Winter 2025", "Summer 2026")) %>%
  mutate(wave = factor(as.character(wave), levels = c("Winter 2025", "Summer 2026")))

wide_common <- atlas_common %>%
  select(school, treatment, student_id, subject, wave, score) %>%
  pivot_wider(names_from = wave, values_from = score) %>%
  filter(!is.na(`Winter 2025`), !is.na(`Summer 2026`)) %>%
  mutate(gain = `Summer 2026` - `Winter 2025`)

kable(head(atlas_long, 10), caption = "First 10 standardized ATLAS observations")
```

# Data availability

```{r data-availability}
data_availability <- atlas_long %>%
  count(school, subject, wave, name = "n_scores") %>%
  arrange(subject, school, wave)

kable(data_availability, caption = "Available score counts by school, subject, and wave")
```

# Mean score for each wave

```{r mean-scores}
mean_scores <- atlas_long %>%
  group_by(school, treatment, subject, wave) %>%
  summarise(
    n = n(),
    mean_score = mean(score, na.rm = TRUE),
    sd_score = sd(score, na.rm = TRUE),
    se_score = sd_score / sqrt(n),
    .groups = "drop"
  ) %>%
  arrange(subject, wave, school)

kable(mean_scores, digits = 2, caption = "Mean Kindergarten ATLAS scores by school, subject, and wave")
```

# Line graph by subject

Each plot shows the mean ATLAS score by wave for both schools. JES Summer 2025 is plotted when available; JMTES begins at Winter 2025 because Summer 2025 is not available in the JMTES Kindergarten workbook.

```{r line-graphs, fig.height=6.5}
mean_line_plot <- ggplot(mean_scores, aes(x = wave, y = mean_score, color = school, group = school)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_score - 1.96 * se_score, ymax = mean_score + 1.96 * se_score), width = 0.08) +
  facet_wrap(~ subject, scales = "free_y") +
  labs(
    title = "Mean Kindergarten ATLAS Scores by School and Wave",
    subtitle = "JMTES received G.R.O.W; JES is the control school",
    x = "Assessment wave",
    y = "Mean ATLAS score",
    color = "School"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

mean_line_plot

dir.create("results", showWarnings = FALSE)
ggsave("results/kindergarten_atlas_mean_scores_line_plot.png", mean_line_plot, width = 9, height = 6.5, dpi = 300)
```

# Primary impact analysis: ANCOVA

## Why ANCOVA is the preferred primary analysis

For the current Kindergarten ATLAS data, **ANCOVA is the preferred primary analysis** because there is only one common pre-period for JMTES and JES. A causal difference-in-differences design depends on a parallel trends assumption, but that assumption cannot be evaluated with only one common pre-treatment ATLAS scale-score wave. ANCOVA instead estimates the difference in Summer 2026 scores between JMTES and JES while controlling for each student's Winter 2025 score.

The ANCOVA model is estimated separately by subject:

\[
Summer2026_i = \alpha + \beta_1 Treatment_i + \beta_2 Winter2025_i + \epsilon_i
\]

where:

- \(Treatment_i = 1\) for JMTES students and 0 for JES students.
- \(Winter2025_i\) is the student's baseline ATLAS score in the same subject.
- \(\beta_1\) is the adjusted JMTES-vs-JES difference in Summer 2026 scores.

A positive and statistically significant \(\beta_1\) would indicate that JMTES students scored higher than comparable JES students in Summer 2026 after adjusting for Winter 2025 baseline performance. This is still quasi-experimental, but it is a better primary model than DiD for these data because it directly adjusts for baseline achievement without requiring an untestable parallel trends assumption.

```{r primary-ancova}
run_ancova <- function(data) {
  lm(`Summer 2026` ~ treatment + `Winter 2025`, data = data)
}

ancova_models <- wide_common %>%
  group_by(subject) %>%
  group_split() %>%
  setNames(levels(droplevels(wide_common$subject))) %>%
  lapply(run_ancova)

ancova_results <- bind_rows(lapply(names(ancova_models), function(s) {
  model <- ancova_models[[s]]
  tidy(model, conf.int = TRUE) %>%
    mutate(subject = s, .before = 1)
}))

ancova_effects <- ancova_results %>%
  filter(term == "treatment") %>%
  transmute(
    subject,
    ancova_estimate = estimate,
    std_error = std.error,
    statistic,
    p_value = p.value,
    conf_low = conf.low,
    conf_high = conf.high
  )

kable(ancova_results, digits = 3, caption = "Primary ANCOVA results by subject")
kable(ancova_effects, digits = 3, caption = "Adjusted JMTES-vs-JES difference from primary ANCOVA model")
```

# Sensitivity check: Difference-in-differences-style estimate

## Why include a DiD-style comparison, and what are its limits?

With the current Kindergarten files, **difference-in-differences should not be described as the best causal option in a strong sense**, because the key DiD identification assumption -- parallel trends -- cannot be tested with only one common pre-period. JMTES and JES share a Winter 2025 pre-period and Summer 2026 post-period, but there are not multiple comparable pre-G.R.O.W ATLAS scale-score waves for both schools that would show whether the schools were trending similarly before G.R.O.W.

The DiD-style estimate is still useful as a transparent sensitivity check because:

1. JMTES received G.R.O.W and JES did not.
2. Both schools have a common pre-period, Winter 2025.
3. Both schools have a common post-period, Summer 2026.
4. The evaluation question is whether JMTES improved more than JES over the same time interval.

However, the estimate should be interpreted as a **descriptive/quasi-experimental pre/post comparison**, not as definitive causal evidence. A stronger causal DiD design would require additional pre-treatment waves for both schools, evidence that pre-treatment ATLAS trends were parallel, or additional comparable control schools. For this reason, this notebook uses ANCOVA as the primary model and reports DiD-style results as a sensitivity check rather than relying on DiD as the main estimate.

The DiD-style estimate is the interaction term from this model, estimated separately by subject:

\[
Score_{ist} = \alpha + \beta_1 Treatment_s + \beta_2 Post_t + \beta_3 (Treatment_s \times Post_t) + \epsilon_{ist}
\]

where:

- \(Treatment_s = 1\) for JMTES and 0 for JES.
- \(Post_t = 1\) for Summer 2026 and 0 for Winter 2025.
- \(\beta_3\) is the DiD-style estimate of the difference in score change between JMTES and JES.

A positive and statistically significant \(\beta_3\) would show that JMTES improved more than JES over the observed period. Without evidence supporting parallel trends, that pattern is consistent with a possible G.R.O.W effect but should not be interpreted as conclusive causal proof.

```{r did-models}
did_data <- atlas_common %>%
  mutate(
    post = if_else(wave == "Summer 2026", 1L, 0L),
    treatment_post = treatment * post
  )

run_did <- function(data) {
  lm(score ~ treatment + post + treatment_post, data = data)
}

did_models <- did_data %>%
  group_by(subject) %>%
  group_split() %>%
  setNames(levels(droplevels(did_data$subject))) %>%
  lapply(run_did)

did_results <- bind_rows(lapply(names(did_models), function(s) {
  model <- did_models[[s]]
  tidy(model, conf.int = TRUE) %>%
    mutate(subject = s, .before = 1)
}))

kable(did_results, digits = 3, caption = "Difference-in-differences regression results by subject")
```

## DiD sensitivity estimate only

```{r did-estimate-only}
did_effects <- did_results %>%
  filter(term == "treatment_post") %>%
  transmute(
    subject,
    did_estimate = estimate,
    std_error = std.error,
    statistic,
    p_value = p.value,
    conf_low = conf.low,
    conf_high = conf.high
  )

kable(did_effects, digits = 3, caption = "DiD-style sensitivity estimate")
```

## Manual DiD-style sensitivity calculation from group means

This table calculates the same estimate directly from mean scores:

\[
(JMTES_{post} - JMTES_{pre}) - (JES_{post} - JES_{pre})
\]

```{r manual-did}
manual_did <- mean_scores %>%
  filter(wave %in% c("Winter 2025", "Summer 2026")) %>%
  select(school, subject, wave, mean_score) %>%
  pivot_wider(names_from = c(school, wave), values_from = mean_score) %>%
  mutate(
    jmtes_change = `JMTES_Summer 2026` - `JMTES_Winter 2025`,
    jes_change = `JES_Summer 2026` - `JES_Winter 2025`,
    did = jmtes_change - jes_change
  )

kable(manual_did, digits = 2, caption = "Manual DiD-style sensitivity calculation from mean scores")
```

# Robustness checks

Because the evaluation has one treatment school and one control school, results should be interpreted cautiously. The checks below test whether the conclusion is sensitive to model specification.

## Robustness check 1: Gain-score model

This model compares mean student-level gains from Winter 2025 to Summer 2026 between the treatment and control schools.

```{r gain-score}
gain_models <- wide_common %>%
  group_by(subject) %>%
  group_split() %>%
  setNames(levels(droplevels(wide_common$subject))) %>%
  lapply(function(data) lm(gain ~ treatment, data = data))

gain_results <- bind_rows(lapply(names(gain_models), function(s) {
  model <- gain_models[[s]]
  tidy(model, conf.int = TRUE) %>%
    mutate(subject = s, .before = 1)
}))

gain_summary <- wide_common %>%
  group_by(subject, school) %>%
  summarise(
    n = n(),
    mean_gain = mean(gain, na.rm = TRUE),
    sd_gain = sd(gain, na.rm = TRUE),
    .groups = "drop"
  )

kable(gain_summary, digits = 2, caption = "Student-level gain summaries")
kable(gain_results, digits = 3, caption = "Gain-score robustness model by subject")
```

## Robustness check 2: Heteroskedasticity-robust standard errors for DiD-style model

This check keeps the same DiD specification but uses HC3 robust standard errors.

```{r robust-se}
robust_did_results <- bind_rows(lapply(names(did_models), function(s) {
  model <- did_models[[s]]
  robust_test <- coeftest(model, vcov. = vcovHC(model, type = "HC3"))
  tibble(
    subject = s,
    term = rownames(robust_test),
    estimate = robust_test[, 1],
    robust_se = robust_test[, 2],
    statistic = robust_test[, 3],
    p_value = robust_test[, 4]
  )
}))

kable(robust_did_results, digits = 3, caption = "DiD model with HC3 robust standard errors")
```

## Robustness check 3: Non-parametric Wilcoxon test on gains

This test compares the distribution of student-level gains between schools without assuming normally distributed gains.

```{r wilcoxon-gains}
wilcoxon_results <- wide_common %>%
  group_by(subject) %>%
  summarise(
    n_jes = sum(school == "JES"),
    n_jmtes = sum(school == "JMTES"),
    wilcox_p_value = wilcox.test(gain ~ school, exact = FALSE)$p.value,
    .groups = "drop"
  )

kable(wilcoxon_results, digits = 3, caption = "Wilcoxon rank-sum test comparing gains by school")
```

## Robustness check 4: Winsorized gain-score comparison

This check limits the influence of extreme gain scores by winsorizing gains at the 5th and 95th percentiles within subject.

```{r winsorized-gains}
winsorize <- function(x, probs = c(0.05, 0.95)) {
  bounds <- quantile(x, probs = probs, na.rm = TRUE)
  pmin(pmax(x, bounds[[1]]), bounds[[2]])
}

wide_winsor <- wide_common %>%
  group_by(subject) %>%
  mutate(gain_winsor = winsorize(gain)) %>%
  ungroup()

winsor_models <- wide_winsor %>%
  group_by(subject) %>%
  group_split() %>%
  setNames(levels(droplevels(wide_winsor$subject))) %>%
  lapply(function(data) lm(gain_winsor ~ treatment, data = data))

winsor_results <- bind_rows(lapply(names(winsor_models), function(s) {
  model <- winsor_models[[s]]
  tidy(model, conf.int = TRUE) %>%
    mutate(subject = s, .before = 1)
}))

kable(winsor_results, digits = 3, caption = "Winsorized gain-score robustness model by subject")
```


## Robustness check 5: Preferred matching method: calipered nearest-neighbor matching on baseline score

Matching can work here as a sensitivity analysis because both schools have student-level Winter 2025 baseline scores. With the current Kindergarten files, the strongest common pre-treatment covariate is the Winter 2025 ATLAS score. Because there is only one strong shared covariate, propensity score matching would be nearly equivalent to matching directly on Winter 2025 score. The preferred matching approach for these files is therefore:

1. **Exact matching by subject**: ELA students are only matched to ELA students, and Math students are only matched to Math students.
2. **Nearest-neighbor matching on Winter 2025 baseline score**: each JMTES student is matched to the JES student with the closest Winter 2025 score.
3. **Matching with replacement**: the same JES student may be reused if that student is the closest match for more than one JMTES student. This improves match quality when the control group is limited.
4. **Caliper restriction**: matches are only accepted when the baseline-score distance is within 0.20 pooled standard deviations of Winter 2025 scores for that subject. This avoids poor matches.

Important limitation: this matching method only adjusts for observed baseline score differences. It cannot adjust for unobserved differences between schools, classrooms, families, or students. ANCOVA remains the primary analysis; matching is a robustness check.

```{r calipered-nearest-neighbor-matching}
pooled_sd <- function(x_treated, x_control) {
  n_t <- length(x_treated)
  n_c <- length(x_control)
  sqrt(((n_t - 1) * var(x_treated, na.rm = TRUE) + (n_c - 1) * var(x_control, na.rm = TRUE)) / (n_t + n_c - 2))
}

caliper_match_subject <- function(data, caliper_sd = 0.20) {
  treated <- data %>% filter(treatment == 1) %>% arrange(student_id)
  controls <- data %>% filter(treatment == 0) %>% arrange(student_id)

  caliper_points <- caliper_sd * pooled_sd(treated$`Winter 2025`, controls$`Winter 2025`)

  bind_rows(lapply(seq_len(nrow(treated)), function(i) {
    baseline_distance <- abs(controls$`Winter 2025` - treated$`Winter 2025`[i])
    match_index <- which.min(baseline_distance)
    accepted_match <- baseline_distance[match_index] <= caliper_points

    tibble(
      subject = as.character(treated$subject[i]),
      treated_id = treated$student_id[i],
      control_id = if_else(accepted_match, controls$student_id[match_index], NA_character_),
      treated_pre = treated$`Winter 2025`[i],
      control_pre = if_else(accepted_match, controls$`Winter 2025`[match_index], NA_real_),
      treated_post = treated$`Summer 2026`[i],
      control_post = if_else(accepted_match, controls$`Summer 2026`[match_index], NA_real_),
      caliper_points = caliper_points,
      baseline_distance = baseline_distance[match_index],
      matched = accepted_match,
      matched_effect = if_else(accepted_match, treated$`Summer 2026`[i] - controls$`Summer 2026`[match_index], NA_real_)
    )
  }))
}

matched_pairs <- wide_common %>%
  group_by(subject) %>%
  group_split() %>%
  lapply(caliper_match_subject) %>%
  bind_rows()

matching_results <- matched_pairs %>%
  group_by(subject) %>%
  summarise(
    n_treated_total = n(),
    n_matched = sum(matched),
    n_unmatched = sum(!matched),
    caliper_points = first(caliper_points),
    mean_baseline_distance = mean(baseline_distance[matched], na.rm = TRUE),
    matched_effect = mean(matched_effect, na.rm = TRUE),
    sd_effect = sd(matched_effect, na.rm = TRUE),
    se_effect = sd_effect / sqrt(n_matched),
    statistic = matched_effect / se_effect,
    p_value = 2 * pt(-abs(statistic), df = n_matched - 1),
    conf_low = matched_effect + qt(0.025, df = n_matched - 1) * se_effect,
    conf_high = matched_effect + qt(0.975, df = n_matched - 1) * se_effect,
    .groups = "drop"
  )

matching_balance <- matched_pairs %>%
  filter(matched) %>%
  group_by(subject) %>%
  summarise(
    treated_pre_mean = mean(treated_pre, na.rm = TRUE),
    matched_control_pre_mean = mean(control_pre, na.rm = TRUE),
    mean_baseline_difference_after_matching = treated_pre_mean - matched_control_pre_mean,
    mean_absolute_baseline_distance = mean(abs(treated_pre - control_pre), na.rm = TRUE),
    .groups = "drop"
  )

kable(matching_results, digits = 3, caption = "Preferred calipered nearest-neighbor baseline-score matching results")
kable(matching_balance, digits = 3, caption = "Baseline balance after preferred matching")
```

## Robustness check 6: Inverse probability weighting using baseline score

Inverse probability weighting (IPW) is another way to adjust for baseline differences between the treatment and control groups. This analysis estimates the probability of being in JMTES as a function of Winter 2025 baseline score, then weights students by the inverse of that probability. With only one baseline covariate, IPW should be treated as exploratory and used to check whether conclusions change after reweighting students to improve baseline-score balance.

```{r ipw}
run_ipw <- function(data) {
  propensity_model <- glm(treatment ~ `Winter 2025`, data = data, family = binomial())

  data_ipw <- data %>%
    mutate(
      propensity = predict(propensity_model, type = "response"),
      propensity = pmin(pmax(propensity, 0.01), 0.99),
      ipw = if_else(treatment == 1, 1 / propensity, 1 / (1 - propensity))
    )

  lm(`Summer 2026` ~ treatment, data = data_ipw, weights = ipw)
}

ipw_models <- wide_common %>%
  group_by(subject) %>%
  group_split() %>%
  setNames(levels(droplevels(wide_common$subject))) %>%
  lapply(run_ipw)

ipw_results <- bind_rows(lapply(names(ipw_models), function(s) {
  model <- ipw_models[[s]]
  tidy(model, conf.int = TRUE) %>%
    mutate(subject = s, .before = 1)
}))

kable(ipw_results, digits = 3, caption = "Inverse probability weighted comparison by subject")
```

## Robustness check 7: Permutation test for the ANCOVA treatment effect

The parametric p-values above rely on regression assumptions. As an additional sensitivity check, this section conducts a student-level permutation test for the ANCOVA treatment coefficient. It repeatedly shuffles treatment labels within subject while preserving the number of JMTES and JES students, then compares the observed ANCOVA treatment coefficient to the permutation distribution.

This is not a perfect school-level randomization test because the intervention occurred at the school level and there is only one treated school and one control school. It is best interpreted as an exploratory check of how unusual the observed adjusted treatment coefficient is under random reassignment of student labels.

```{r permutation-test}
permute_ancova <- function(data, n_perm = 2000, seed = 2026) {
  set.seed(seed)
  observed <- coef(lm(`Summer 2026` ~ treatment + `Winter 2025`, data = data))["treatment"]

  permuted <- replicate(n_perm, {
    data_perm <- data %>%
      mutate(treatment = sample(treatment, size = n(), replace = FALSE))
    coef(lm(`Summer 2026` ~ treatment + `Winter 2025`, data = data_perm))["treatment"]
  })

  tibble(
    observed_effect = unname(observed),
    n_permutations = n_perm,
    permutation_p_value = mean(abs(permuted) >= abs(observed), na.rm = TRUE),
    permutation_mean = mean(permuted, na.rm = TRUE),
    permutation_sd = sd(permuted, na.rm = TRUE)
  )
}

permutation_results <- wide_common %>%
  group_by(subject) %>%
  group_split() %>%
  setNames(levels(droplevels(wide_common$subject))) %>%
  lapply(permute_ancova) %>%
  bind_rows(.id = "subject")

kable(permutation_results, digits = 3, caption = "Permutation test for ANCOVA treatment coefficient")
```

## Robustness check 8: Bootstrap confidence intervals for ANCOVA

This check uses non-parametric bootstrap resampling to estimate uncertainty around the ANCOVA treatment coefficient. It resamples students within each subject and re-estimates the ANCOVA model. Bootstrap intervals are useful because they do not rely as heavily on the standard linear-model standard-error formula.

```{r bootstrap-ancova}
bootstrap_ancova <- function(data, n_boot = 2000, seed = 2026) {
  set.seed(seed)
  observed <- coef(lm(`Summer 2026` ~ treatment + `Winter 2025`, data = data))["treatment"]

  boot_estimates <- replicate(n_boot, {
    boot_index <- sample(seq_len(nrow(data)), size = nrow(data), replace = TRUE)
    boot_data <- data[boot_index, ]
    coef(lm(`Summer 2026` ~ treatment + `Winter 2025`, data = boot_data))["treatment"]
  })

  tibble(
    observed_effect = unname(observed),
    n_bootstrap = n_boot,
    bootstrap_se = sd(boot_estimates, na.rm = TRUE),
    conf_low = quantile(boot_estimates, 0.025, na.rm = TRUE),
    conf_high = quantile(boot_estimates, 0.975, na.rm = TRUE)
  )
}

bootstrap_results <- wide_common %>%
  group_by(subject) %>%
  group_split() %>%
  setNames(levels(droplevels(wide_common$subject))) %>%
  lapply(bootstrap_ancova) %>%
  bind_rows(.id = "subject")

kable(bootstrap_results, digits = 3, caption = "Bootstrap confidence intervals for ANCOVA treatment coefficient")
```

# Other methods considered but not recommended with current data

The following methods are common for impact analysis, but they are not appropriate as primary analyses with the current Kindergarten files:

- **Full causal DiD/event study:** would require multiple comparable pre-G.R.O.W waves for both JMTES and JES to assess parallel trends.
- **Student fixed effects with multiple periods:** would require more than two comparable waves and ideally more pre/post observations.
- **School fixed effects with clustered standard errors:** would require more schools; with one treatment school and one control school, school-level clustering is not meaningful.
- **Regression discontinuity:** would require a known assignment cutoff for G.R.O.W eligibility or participation.
- **Instrumental variables:** would require a valid instrument that affects G.R.O.W participation but affects ATLAS scores only through G.R.O.W.
- **Synthetic control:** would require several control schools and multiple pre-treatment waves to construct a credible synthetic comparison school.

# Interpretation guide

Use the following hierarchy when interpreting the findings:

1. **Primary estimate:** The ANCOVA treatment coefficient, which compares Summer 2026 scores between JMTES and JES after adjusting for Winter 2025 baseline scores.
2. **Substantive magnitude:** Whether the estimated effect is meaningful in ATLAS score points, not only whether it is statistically significant.
3. **Consistency across robustness checks:** Whether the DiD-style, gain-score, matching, IPW, permutation, bootstrap, robust-SE, Wilcoxon, and winsorized analyses point in the same direction as ANCOVA.
4. **Limitations:** Results are quasi-experimental because students were not randomly assigned to G.R.O.W, and there is only one treatment school and one control school.

A positive ANCOVA treatment estimate means JMTES had higher Summer 2026 scores than JES after adjusting for Winter 2025 baseline scores. A positive DiD-style or gain-score estimate means JMTES improved more than JES between Winter 2025 and Summer 2026.

# Recommended reporting language

If the ANCOVA estimate is positive and robustness checks are consistent:

> After adjusting for Winter 2025 baseline scores, JMTES students had higher Summer 2026 ATLAS scores than comparable JES students, which is consistent with a positive G.R.O.W effect. However, because this is a quasi-experimental comparison involving one treatment school and one control school, the estimate should be interpreted as suggestive evidence rather than definitive causal proof.

If the ANCOVA estimate is near zero or inconsistent across checks:

> The Kindergarten ATLAS data do not provide consistent evidence that JMTES had higher Summer 2026 scores than JES after adjusting for Winter 2025 baseline scores. Additional waves, additional comparison schools, and student-level covariates would strengthen the impact analysis.

# Save analytic outputs

```{r save-outputs}
dir.create("results", showWarnings = FALSE)
write.csv(atlas_long, "results/kindergarten_atlas_long.csv", row.names = FALSE)
write.csv(mean_scores, "results/kindergarten_atlas_mean_scores.csv", row.names = FALSE)
write.csv(ancova_effects, "results/kindergarten_atlas_ancova_effects.csv", row.names = FALSE)
write.csv(did_effects, "results/kindergarten_atlas_did_effects.csv", row.names = FALSE)
write.csv(gain_summary, "results/kindergarten_atlas_gain_summary.csv", row.names = FALSE)
write.csv(matched_pairs, "results/kindergarten_atlas_matching_pairs.csv", row.names = FALSE)
write.csv(matching_results, "results/kindergarten_atlas_matching_results.csv", row.names = FALSE)
write.csv(matching_balance, "results/kindergarten_atlas_matching_balance.csv", row.names = FALSE)
write.csv(ipw_results, "results/kindergarten_atlas_ipw_results.csv", row.names = FALSE)
write.csv(permutation_results, "results/kindergarten_atlas_permutation_results.csv", row.names = FALSE)
write.csv(bootstrap_results, "results/kindergarten_atlas_bootstrap_results.csv", row.names = FALSE)
```
