---
title: "Kindergarten ATLAS Assessment Analysis: G.R.O.W Impact at JMTES"
author: "CSIC Evaluation Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5.5
)
```

# Purpose

This notebook analyzes Kindergarten ATLAS assessment scores for the G.R.O.W evaluation.

- **Treatment school:** JMTES / Murrell Taylor Elementary, which received G.R.O.W.
- **Control school:** JES, which did not receive G.R.O.W.
- **Primary outcome:** ATLAS assessment scale scores.
- **Subjects:** ELA and Math.
- **Primary analytic question:** Did Kindergarten ATLAS assessment scores improve more at the treatment school than at the control school after G.R.O.W implementation?

The strongest common pre/post comparison available in the two Kindergarten workbooks is:

- **Pre period:** Winter 2025
- **Post period:** Summer 2026

JES also includes Summer 2025 scores, but the JMTES Kindergarten workbook does not include Summer 2025. Therefore, the impact analysis focuses on the two waves available for both schools.

# Load packages

```{r packages}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(broom)
library(knitr)
library(sandwich)
library(lmtest)
```

# File paths and helper functions

```{r paths-and-helpers}
jmtes_file <- "JMTES Atlas Report - Kindergarten.xlsx"
jes_file   <- "JES Atlas Report - Kindergarten.xlsx"

clean_score <- function(x) {
  suppressWarnings(as.numeric(x))
}

wave_order <- c("Summer 2025", "Winter 2025", "Summer 2026")
subject_order <- c("ELA", "Math")
```

# Import and reshape ATLAS data

The two school files use slightly different column names. This section standardizes the data into one long-format file with one row per student, school, subject, and wave.

```{r import-data}
# JMTES treatment school. Row 1 in this workbook is a title row, so data headers begin on row 2.
jmtes_raw <- read_excel(jmtes_file, skip = 1, col_types = "text")

jmtes_long <- jmtes_raw %>%
  rename(
    student_id = `Student ID`,
    grade = Grade,
    teacher = Teacher
  ) %>%
  mutate(
    school = "JMTES",
    treatment = 1L
  ) %>%
  pivot_longer(
    cols = matches("^(ELA|Math)"),
    names_to = "assessment",
    values_to = "score"
  ) %>%
  mutate(
    score = clean_score(score),
    subject = case_when(
      str_detect(assessment, "^ELA") ~ "ELA",
      str_detect(assessment, "^Math") ~ "Math",
      TRUE ~ NA_character_
    ),
    wave = case_when(
      str_detect(assessment, regex("Winter 2025", ignore_case = TRUE)) ~ "Winter 2025",
      str_detect(assessment, regex("Sum 2026|Summer 2026", ignore_case = TRUE)) ~ "Summer 2026",
      str_detect(assessment, regex("Sum 2025|Summer 2025", ignore_case = TRUE)) ~ "Summer 2025",
      TRUE ~ NA_character_
    )
  ) %>%
  select(school, treatment, grade, teacher, student_id, subject, wave, score)

# JES control school.
jes_raw <- read_excel(jes_file, col_types = "text")

jes_long <- jes_raw %>%
  rename(
    student_id = `ID Number`,
    grade = Grade
  ) %>%
  mutate(
    school = "JES",
    treatment = 0L,
    teacher = NA_character_
  ) %>%
  pivot_longer(
    cols = matches("^(ELA|Math)"),
    names_to = "assessment",
    values_to = "score"
  ) %>%
  mutate(
    score = clean_score(score),
    subject = case_when(
      str_detect(assessment, "^ELA") ~ "ELA",
      str_detect(assessment, "^Math") ~ "Math",
      TRUE ~ NA_character_
    ),
    wave = case_when(
      str_detect(assessment, regex("Winter 2025", ignore_case = TRUE)) ~ "Winter 2025",
      str_detect(assessment, regex("Sum 2026|Summer 2026", ignore_case = TRUE)) ~ "Summer 2026",
      str_detect(assessment, regex("Sum 2025|Summer 2025", ignore_case = TRUE)) ~ "Summer 2025",
      TRUE ~ NA_character_
    )
  ) %>%
  select(school, treatment, grade, teacher, student_id, subject, wave, score)

atlas_long <- bind_rows(jmtes_long, jes_long) %>%
  filter(!is.na(subject), !is.na(wave), !is.na(score)) %>%
  mutate(
    school = factor(school, levels = c("JES", "JMTES")),
    treatment = as.integer(treatment),
    subject = factor(subject, levels = subject_order),
    wave = factor(wave, levels = wave_order),
    post = if_else(wave == "Summer 2026", 1L, 0L)
  )

atlas_common <- atlas_long %>%
  filter(wave %in% c("Winter 2025", "Summer 2026")) %>%
  mutate(wave = factor(as.character(wave), levels = c("Winter 2025", "Summer 2026")))

kable(head(atlas_long, 10), caption = "First 10 standardized ATLAS observations")
```

# Data availability

```{r data-availability}
data_availability <- atlas_long %>%
  count(school, subject, wave, name = "n_scores") %>%
  arrange(subject, school, wave)

kable(data_availability, caption = "Available score counts by school, subject, and wave")
```

# Mean score for each wave

```{r mean-scores}
mean_scores <- atlas_long %>%
  group_by(school, treatment, subject, wave) %>%
  summarise(
    n = n(),
    mean_score = mean(score, na.rm = TRUE),
    sd_score = sd(score, na.rm = TRUE),
    se_score = sd_score / sqrt(n),
    .groups = "drop"
  ) %>%
  arrange(subject, wave, school)

kable(mean_scores, digits = 2, caption = "Mean Kindergarten ATLAS scores by school, subject, and wave")
```

# Line graph by subject

Each plot shows the mean ATLAS score by wave for both schools. JES Summer 2025 is plotted when available; JMTES begins at Winter 2025 because Summer 2025 is not available in the JMTES Kindergarten workbook.

```{r line-graphs, fig.height=6.5}
ggplot(mean_scores, aes(x = wave, y = mean_score, color = school, group = school)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_score - 1.96 * se_score, ymax = mean_score + 1.96 * se_score), width = 0.08) +
  facet_wrap(~ subject, scales = "free_y") +
  labs(
    title = "Mean Kindergarten ATLAS Scores by School and Wave",
    subtitle = "JMTES received G.R.O.W; JES is the control school",
    x = "Assessment wave",
    y = "Mean ATLAS score",
    color = "School"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

# Primary impact analysis: Difference-in-differences

## Why difference-in-differences?

A difference-in-differences (DiD) design is the most appropriate first impact analysis for these data because:

1. JMTES received G.R.O.W and JES did not.
2. Both schools have a common pre period, Winter 2025.
3. Both schools have a common post period, Summer 2026.
4. The key question is whether JMTES improved more than JES over the same time interval.

The DiD estimate is the interaction term from this model, estimated separately by subject:

\[
Score_{ist} = \alpha + \beta_1 Treatment_s + \beta_2 Post_t + \beta_3 (Treatment_s \times Post_t) + \epsilon_{ist}
\]

where:

- \(Treatment_s = 1\) for JMTES and 0 for JES.
- \(Post_t = 1\) for Summer 2026 and 0 for Winter 2025.
- \(\beta_3\) is the estimated G.R.O.W impact on ATLAS scores.

A positive and statistically significant \(\beta_3\) would suggest that JMTES improved more than JES after G.R.O.W implementation. This is still a quasi-experimental estimate, not a randomized causal estimate.

```{r did-models}
did_data <- atlas_common %>%
  mutate(
    post = if_else(wave == "Summer 2026", 1L, 0L),
    treatment_post = treatment * post
  )

run_did <- function(data) {
  lm(score ~ treatment + post + treatment_post, data = data)
}

did_models <- did_data %>%
  group_by(subject) %>%
  group_split() %>%
  setNames(levels(droplevels(did_data$subject))) %>%
  lapply(run_did)

did_results <- bind_rows(lapply(names(did_models), function(s) {
  model <- did_models[[s]]
  tidy(model, conf.int = TRUE) %>%
    mutate(subject = s, .before = 1)
}))

kable(did_results, digits = 3, caption = "Difference-in-differences regression results by subject")
```

## DiD estimate only

```{r did-estimate-only}
did_effects <- did_results %>%
  filter(term == "treatment_post") %>%
  transmute(
    subject,
    did_estimate = estimate,
    std_error = std.error,
    statistic,
    p_value = p.value,
    conf_low = conf.low,
    conf_high = conf.high
  )

kable(did_effects, digits = 3, caption = "Estimated G.R.O.W impact from DiD model")
```

## Manual DiD calculation from group means

This table calculates the same estimate directly from mean scores:

\[
(JMTES_{post} - JMTES_{pre}) - (JES_{post} - JES_{pre})
\]

```{r manual-did}
manual_did <- mean_scores %>%
  filter(wave %in% c("Winter 2025", "Summer 2026")) %>%
  select(school, subject, wave, mean_score) %>%
  pivot_wider(names_from = c(school, wave), values_from = mean_score) %>%
  mutate(
    jmtes_change = `JMTES_Summer 2026` - `JMTES_Winter 2025`,
    jes_change = `JES_Summer 2026` - `JES_Winter 2025`,
    did = jmtes_change - jes_change
  )

kable(manual_did, digits = 2, caption = "Manual difference-in-differences calculation from mean scores")
```

# Robustness checks

Because the evaluation has one treatment school and one control school, results should be interpreted cautiously. The checks below test whether the conclusion is sensitive to model specification.

## Robustness check 1: ANCOVA model

ANCOVA compares Summer 2026 scores between schools while controlling for each student's Winter 2025 score. This is often more statistically efficient than a simple gain-score comparison when student-level pretest scores are available.

```{r ancova}
wide_common <- atlas_common %>%
  select(school, treatment, student_id, subject, wave, score) %>%
  pivot_wider(names_from = wave, values_from = score) %>%
  filter(!is.na(`Winter 2025`), !is.na(`Summer 2026`)) %>%
  mutate(gain = `Summer 2026` - `Winter 2025`)

run_ancova <- function(data) {
  lm(`Summer 2026` ~ treatment + `Winter 2025`, data = data)
}

ancova_models <- wide_common %>%
  group_by(subject) %>%
  group_split() %>%
  setNames(levels(droplevels(wide_common$subject))) %>%
  lapply(run_ancova)

ancova_results <- bind_rows(lapply(names(ancova_models), function(s) {
  model <- ancova_models[[s]]
  tidy(model, conf.int = TRUE) %>%
    mutate(subject = s, .before = 1)
}))

kable(ancova_results, digits = 3, caption = "ANCOVA robustness check by subject")
```

## Robustness check 2: Gain-score model

This model compares mean student-level gains from Winter 2025 to Summer 2026 between the treatment and control schools.

```{r gain-score}
gain_models <- wide_common %>%
  group_by(subject) %>%
  group_split() %>%
  setNames(levels(droplevels(wide_common$subject))) %>%
  lapply(function(data) lm(gain ~ treatment, data = data))

gain_results <- bind_rows(lapply(names(gain_models), function(s) {
  model <- gain_models[[s]]
  tidy(model, conf.int = TRUE) %>%
    mutate(subject = s, .before = 1)
}))

gain_summary <- wide_common %>%
  group_by(subject, school) %>%
  summarise(
    n = n(),
    mean_gain = mean(gain, na.rm = TRUE),
    sd_gain = sd(gain, na.rm = TRUE),
    .groups = "drop"
  )

kable(gain_summary, digits = 2, caption = "Student-level gain summaries")
kable(gain_results, digits = 3, caption = "Gain-score robustness model by subject")
```

## Robustness check 3: Heteroskedasticity-robust standard errors

This check keeps the same DiD specification but uses HC3 robust standard errors.

```{r robust-se}
robust_did_results <- bind_rows(lapply(names(did_models), function(s) {
  model <- did_models[[s]]
  robust_test <- coeftest(model, vcov. = vcovHC(model, type = "HC3"))
  tibble(
    subject = s,
    term = rownames(robust_test),
    estimate = robust_test[, 1],
    robust_se = robust_test[, 2],
    statistic = robust_test[, 3],
    p_value = robust_test[, 4]
  )
}))

kable(robust_did_results, digits = 3, caption = "DiD model with HC3 robust standard errors")
```

## Robustness check 4: Non-parametric Wilcoxon test on gains

This test compares the distribution of student-level gains between schools without assuming normally distributed gains.

```{r wilcoxon-gains}
wilcoxon_results <- wide_common %>%
  group_by(subject) %>%
  summarise(
    n_jes = sum(school == "JES"),
    n_jmtes = sum(school == "JMTES"),
    wilcox_p_value = wilcox.test(gain ~ school, exact = FALSE)$p.value,
    .groups = "drop"
  )

kable(wilcoxon_results, digits = 3, caption = "Wilcoxon rank-sum test comparing gains by school")
```

## Robustness check 5: Winsorized gain-score comparison

This check limits the influence of extreme gain scores by winsorizing gains at the 5th and 95th percentiles within subject.

```{r winsorized-gains}
winsorize <- function(x, probs = c(0.05, 0.95)) {
  bounds <- quantile(x, probs = probs, na.rm = TRUE)
  pmin(pmax(x, bounds[[1]]), bounds[[2]])
}

wide_winsor <- wide_common %>%
  group_by(subject) %>%
  mutate(gain_winsor = winsorize(gain)) %>%
  ungroup()

winsor_models <- wide_winsor %>%
  group_by(subject) %>%
  group_split() %>%
  setNames(levels(droplevels(wide_winsor$subject))) %>%
  lapply(function(data) lm(gain_winsor ~ treatment, data = data))

winsor_results <- bind_rows(lapply(names(winsor_models), function(s) {
  model <- winsor_models[[s]]
  tidy(model, conf.int = TRUE) %>%
    mutate(subject = s, .before = 1)
}))

kable(winsor_results, digits = 3, caption = "Winsorized gain-score robustness model by subject")
```

# Interpretation guide

Use the following hierarchy when interpreting the findings:

1. **Primary estimate:** The DiD interaction term, `treatment_post`.
2. **Substantive magnitude:** Whether the estimated effect is meaningful in ATLAS score points, not only whether it is statistically significant.
3. **Consistency across robustness checks:** Whether the ANCOVA, gain-score, robust-SE, Wilcoxon, and winsorized analyses point in the same direction.
4. **Limitations:** Results are quasi-experimental because students were not randomly assigned to G.R.O.W, and there is only one treatment school and one control school.

A positive DiD estimate means that JMTES improved more than JES between Winter 2025 and Summer 2026. A negative DiD estimate means that JMTES improved less than JES over the same period.

# Recommended reporting language

If the DiD estimate is positive and robustness checks are consistent:

> Kindergarten ATLAS scores increased more at JMTES than at JES between Winter 2025 and Summer 2026, which is consistent with a positive G.R.O.W impact. However, because this is a quasi-experimental comparison involving one treatment school and one control school, the estimate should be interpreted as suggestive evidence rather than definitive causal proof.

If the DiD estimate is near zero or inconsistent across checks:

> The Kindergarten ATLAS data do not provide consistent evidence that JMTES improved more than JES between Winter 2025 and Summer 2026. Additional waves, additional comparison schools, and student-level covariates would strengthen the impact analysis.

# Save analytic outputs

```{r save-outputs}
write.csv(atlas_long, "kindergarten_atlas_long.csv", row.names = FALSE)
write.csv(mean_scores, "kindergarten_atlas_mean_scores.csv", row.names = FALSE)
write.csv(did_effects, "kindergarten_atlas_did_effects.csv", row.names = FALSE)
write.csv(gain_summary, "kindergarten_atlas_gain_summary.csv", row.names = FALSE)
```
