---
title: "2026 Summative ATLAS Descriptive Statistics"
output:
  word_document:
    reference_docx: null
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(dplyr)
library(readr)
library(readxl)
library(knitr)
```

```{r helper-functions}
grade_folders <- c("Kindergarten", "Grade 1", "Grade 2", "Grade 3", "Grade 4", "Grade 5")
subjects_by_grade <- list(
  "Kindergarten" = c("ELA", "Math"),
  "Grade 1" = c("ELA", "Math"),
  "Grade 2" = c("ELA", "Math"),
  "Grade 3" = c("ELA", "Math", "Science"),
  "Grade 4" = c("ELA", "Math", "Science"),
  "Grade 5" = c("ELA", "Math", "Science")
)

find_atlas_file <- function(grade, school) {
  files <- list.files(grade, pattern = paste0("^", school, ".*Atlas Report.*\\.xlsx$"), full.names = TRUE)
  current_files <- files[grepl("Current Year", files)]
  if (length(current_files) > 0) return(current_files[[1]])
  files[[1]]
}

read_atlas_file <- function(path) {
  raw <- read_excel(path, col_names = FALSE)
  header_row <- which(apply(raw, 1, function(x) any(trimws(as.character(x)) %in% c("Student ID", "Student IDs", "ID Number"))))[[1]]
  names <- as.character(unlist(raw[header_row, ]))
  dat <- raw[(header_row + 1):nrow(raw), ]
  colnames(dat) <- names
  dat
}

score_values <- function(dat, subject) {
  score_col <- paste(subject, "Sum 2026")
  as.numeric(dat[[score_col]])
}

summarize_scores <- function(values) {
  values <- values[!is.na(values)]
  tibble(
    n = length(values),
    mean = mean(values),
    sd = sd(values),
    median = median(values)
  )
}
```

```{r calculate-table}
atlas_stats <- bind_rows(lapply(grade_folders, function(grade) {
  jmtes <- read_atlas_file(find_atlas_file(grade, "JMTES"))
  jes <- read_atlas_file(find_atlas_file(grade, "JES"))

  bind_rows(lapply(subjects_by_grade[[grade]], function(subject) {
    jmtes_stats <- summarize_scores(score_values(jmtes, subject))
    jes_stats <- summarize_scores(score_values(jes, subject))

    tibble(
      Grade = grade,
      Subject = subject,
      `JMTES\n mean\n (SD)` = sprintf("%.2f\n (%.2f)", jmtes_stats$mean, jmtes_stats$sd),
      `JES\n mean\n (SD)` = sprintf("%.2f\n (%.2f)", jes_stats$mean, jes_stats$sd),
      `Difference-in-mean\n (JMTES - JES)` = sprintf("%.2f", jmtes_stats$mean - jes_stats$mean),
      `JMTES\n median` = format(jmtes_stats$median, trim = TRUE, nsmall = 0),
      `JES\n median` = format(jes_stats$median, trim = TRUE, nsmall = 0),
      `Observations\n (JMTES/JES)` = paste0(jmtes_stats$n, "/", jes_stats$n)
    )
  }))
}))

write_csv(atlas_stats, "results/atlas_2026_summative_descriptive_statistics.csv")
atlas_stats
```

```{r display-table, results='asis'}
kable(
  atlas_stats,
  format = "pipe",
  col.names = c(
    "Grade", "Subject", "JMTES\n mean\n (SD)", "JES\n mean\n (SD)",
    "Difference-in-mean\n (JMTES - JES)", "JMTES\n median", "JES\n median",
    "Observations\n (JMTES/JES)"
  )
)
```

Notes: The table reports unweighted descriptive statistics for ATLAS assessment scores by grade, subject, and school. Standard deviations are reported in parentheses below means. The difference-in-mean column is JMTES minus JES, so positive values indicate higher average scores at JMTES. Median values are reported separately for each school. Because the number of observations varies across grade-subject cells, observations are reported in the final column as JMTES/JES.
