---
title: "Impact of GROW on Academic Performance: Comparing the ATLAS Assessment Scores Between JMTES and JES"
author: "CSIC Evaluation Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

This R Markdown file is the reproducibility notebook for the Kindergarten through Grade 5 merged at-risk ATLAS impact analysis. It runs the repository analysis driver `../run_atlas_grow_impact.py`, which reads each grade's merged at-risk files and produces three requested regression analyses.

# Analytic Design

The preferred estimate uses an overlap-weighted propensity score approach with a baseline-adjusted and covariate-adjusted outcome regression. The propensity score is estimated from baseline same-subject ATLAS score, gender, race/ethnicity, age, meal-status code, and entry code. JMTES students receive weight `1 - p_i`; JES students receive weight `p_i`. The weighted outcome model regresses 2026 summative ATLAS score on the JMTES indicator, baseline score, gender, race/ethnicity, age, meal status, and entry code.

Two robustness analyses are also produced:

1. **Propensity-score matching plus adjusted regression.** Students are matched using calipered nearest-neighbor matching on the estimated propensity score. The matched sample is then analyzed with a baseline-adjusted and covariate-adjusted regression.
2. **DiD-style gain-score regression.** The dependent variable is the change from baseline ATLAS score to 2026 summative ATLAS score. The model adjusts for baseline score, gender, race/ethnicity, age, meal status, and entry code.

# Run the Analysis

```{r run-analysis}
# This call produces all regression tables, the combined CSV, and the Word-compatible report.
system2("python3", "../run_atlas_grow_impact.py")
```

# Generated Outputs

The analysis creates these primary outputs in this folder:

* `overlap_weighted_ps_aer_regression_table.md`
* `matching_aer_regression_table.md`
* `did_aer_regression_table.md`
* `grow_atlas_impact_regression_results.csv`
* `grow_atlas_impact_results.md`
* `grow_atlas_impact_results.doc`
