---
title: "Impact of GROW on Academic Performance: Comparing the ATLAS Assessment Scores Between JMTES and JES"
author: "CSIC Evaluation Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

This R Markdown file is the reproducibility notebook for the Kindergarten through Grade 5 merged at-risk ATLAS impact analysis. It runs the repository analysis driver `../run_atlas_grow_impact.py`, which reads each grade's merged at-risk files and produces three requested regression analyses.

# Analytic Design

The preferred estimate uses an overlap-weighted propensity score approach with a baseline-adjusted and covariate-adjusted outcome regression. The propensity score is estimated from baseline same-subject ATLAS score, gender, race/ethnicity, age, meal-status code, and entry code. JMTES students receive weight `1 - p_i`; JES students receive weight `p_i`. The weighted outcome model regresses 2026 summative ATLAS score on the JMTES indicator, baseline score, gender, race/ethnicity, age, meal status, and entry code.

Two robustness analyses are also produced:

1. **Propensity-score matching plus adjusted regression.** Students are matched using calipered nearest-neighbor matching on the estimated propensity score. The matched sample is then analyzed with a baseline-adjusted and covariate-adjusted regression.
2. **DiD-style gain-score regression.** The dependent variable is the change from baseline ATLAS score to 2026 summative ATLAS score. The model adjusts for baseline score, gender, race/ethnicity, age, meal status, and entry code.

# Run the Analysis

```{r run-analysis}
# This call produces all regression tables, the combined CSV, and the Word-compatible report.
system2("python3", "../run_atlas_grow_impact.py")
```

# Generated Outputs

The analysis creates these primary outputs in this folder:

* `overlap_weighted_ps_aer_regression_table.md`
* `matching_aer_regression_table.md`
* `did_aer_regression_table.md`
* `grow_atlas_impact_regression_results.csv`
* `grow_atlas_impact_results.md`
* `grow_atlas_impact_results.doc`
* `grow_atlas_preferred_coefficient_plot.png`
* `grow_atlas_matching_coefficient_plot.png`
* `grow_atlas_did_coefficient_plot.png`


# Preferred-Model Coefficient Plot

The coefficient plot below visualizes the preferred overlap-weighted, baseline-adjusted, and covariate-adjusted JMTES coefficients for every available grade-by-subject ATLAS model. Points show the adjusted difference in 2026 Summative ATLAS scale scores between JMTES and comparable JES students; error bars show 95 percent confidence intervals calculated from the model standard errors. Positive values favor JMTES, and the dashed vertical line marks no adjusted difference. Point shapes and coefficient labels use the same significance-star convention as Table 2 (* p < 0.10, ** p < 0.05, *** p < 0.01), so Grade 5 Science is shown as statistically significant just as it is in the table. This visual matches the main results table and discussion: estimates are generally small and not statistically significant in Kindergarten through Grade 2, become consistently positive and statistically significant in Grades 3 and 4, and are mixed in Grade 5, with Science remaining statistically significant.

```{r preferred-coefficient-plot, fig.width=10, fig.height=6, dpi=300}
library(ggplot2)

results_path <- "grow_atlas_impact_regression_results.csv"
if (!file.exists(results_path)) {
  results_path <- file.path("ATLAS GROW Impact Analysis", results_path)
}

preferred_results <- read.csv(results_path, stringsAsFactors = FALSE)
preferred_results <- preferred_results[
  preferred_results$analysis == "overlap_weighted_ps_covariate_baseline_regression",
]

preferred_results$grade <- factor(
  preferred_results$grade,
  levels = c("Kindergarten", "Grade 1", "Grade 2", "Grade 3", "Grade 4", "Grade 5")
)
preferred_results$subject <- factor(
  preferred_results$subject,
  levels = c("ELA", "Math", "Science")
)
preferred_results$ci_lower <- preferred_results$estimate - 1.96 * preferred_results$std_error
preferred_results$ci_upper <- preferred_results$estimate + 1.96 * preferred_results$std_error
preferred_results$significance <- cut(
  preferred_results$p_value,
  breaks = c(-Inf, 0.01, 0.05, 0.10, Inf),
  labels = c("*** p < 0.01", "** p < 0.05", "* p < 0.10", "n.s."),
  right = FALSE
)
preferred_results$stars <- ifelse(
  preferred_results$p_value < 0.01,
  "***",
  ifelse(preferred_results$p_value < 0.05, "**", ifelse(preferred_results$p_value < 0.10, "*", ""))
)
preferred_results$estimate_label <- paste0(sprintf("%.2f", preferred_results$estimate), preferred_results$stars)

coefficient_plot <- ggplot(
  preferred_results,
  aes(x = estimate, y = grade, color = subject, shape = significance)
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray55", linewidth = 0.5) +
  geom_errorbarh(
    aes(xmin = ci_lower, xmax = ci_upper),
    height = 0.18,
    linewidth = 0.75,
    position = position_dodge(width = 0.55)
  ) +
  geom_point(size = 3.2, position = position_dodge(width = 0.55)) +
  geom_text(
    aes(label = estimate_label),
    position = position_dodge(width = 0.55),
    hjust = -0.65,
    size = 3.1,
    show.legend = FALSE
  ) +
  scale_y_discrete(limits = rev(levels(preferred_results$grade))) +
  scale_color_manual(values = c("ELA" = "#1b9e77", "Math" = "#d95f02", "Science" = "#7570b3")) +
  scale_shape_manual(values = c("*** p < 0.01" = 16, "** p < 0.05" = 17, "* p < 0.10" = 15, "n.s." = 1), drop = FALSE) +
  coord_cartesian(xlim = c(-6, 14), clip = "off") +
  labs(
    title = "GROW-Associated Differences in 2026 Summative ATLAS Scores",
    subtitle = "Preferred overlap-weighted, baseline- and covariate-adjusted models; error bars show 95% CIs",
    x = "Adjusted JMTES coefficient in ATLAS scale-score points",
    y = NULL,
    color = "Subject",
    shape = "Table 2 significance"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    plot.margin = margin(10, 35, 10, 10)
  )

coefficient_plot

ggsave(
  filename = file.path(dirname(results_path), "grow_atlas_preferred_coefficient_plot.png"),
  plot = coefficient_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```


# Appendix B Robustness Coefficient Plots

The two coefficient plots below visualize the Appendix B robustness estimates for every available grade-by-subject ATLAS model. The first plot shows the calipered nearest-neighbor matching estimates from Table B1, and the second plot shows the difference-in-differences style gain-score estimates from Table B2. Both plots use the same Table B significance-star convention (* p < 0.10, ** p < 0.05, *** p < 0.01), so the visual labels align with the appendix tables and discussion. In the matching plot, the significant upper-elementary findings include Grade 3 ELA, Grade 3 Math, Grade 3 Science, Grade 4 ELA, Grade 4 Math, Grade 4 Science, Grade 5 ELA, and Grade 5 Science. In the difference-in-differences style plot, the significant findings include Grade 3 ELA, Grade 3 Math, Grade 3 Science, Grade 4 ELA, Grade 4 Math, Grade 4 Science, and Grade 5 Science. Together, the plots reinforce the Appendix B discussion that robustness evidence is strongest in Grades 3 and 4, weaker in Kindergarten through Grade 2, and mixed in Grade 5.

```{r appendix-b-coefficient-plots, fig.width=10, fig.height=6, dpi=300}
library(ggplot2)

results_path <- "grow_atlas_impact_regression_results.csv"
if (!file.exists(results_path)) {
  results_path <- file.path("ATLAS GROW Impact Analysis", results_path)
}

all_results <- read.csv(results_path, stringsAsFactors = FALSE)

prepare_coefficient_data <- function(results, analysis_name) {
  plot_data <- results[results$analysis == analysis_name, ]
  plot_data$grade <- factor(
    plot_data$grade,
    levels = c("Kindergarten", "Grade 1", "Grade 2", "Grade 3", "Grade 4", "Grade 5")
  )
  plot_data$subject <- factor(plot_data$subject, levels = c("ELA", "Math", "Science"))
  plot_data$ci_lower <- plot_data$estimate - 1.96 * plot_data$std_error
  plot_data$ci_upper <- plot_data$estimate + 1.96 * plot_data$std_error
  plot_data$significance <- cut(
    plot_data$p_value,
    breaks = c(-Inf, 0.01, 0.05, 0.10, Inf),
    labels = c("*** p < 0.01", "** p < 0.05", "* p < 0.10", "n.s."),
    right = FALSE
  )
  plot_data$stars <- ifelse(
    plot_data$p_value < 0.01,
    "***",
    ifelse(plot_data$p_value < 0.05, "**", ifelse(plot_data$p_value < 0.10, "*", ""))
  )
  plot_data$estimate_label <- paste0(sprintf("%.2f", plot_data$estimate), plot_data$stars)
  plot_data
}

make_coefficient_plot <- function(plot_data, plot_title, plot_subtitle) {
  ggplot(plot_data, aes(x = estimate, y = grade, color = subject, shape = significance)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray55", linewidth = 0.5) +
    geom_errorbarh(
      aes(xmin = ci_lower, xmax = ci_upper),
      height = 0.18,
      linewidth = 0.75,
      position = position_dodge(width = 0.55)
    ) +
    geom_point(size = 3.2, position = position_dodge(width = 0.55)) +
    geom_text(
      aes(label = estimate_label),
      position = position_dodge(width = 0.55),
      hjust = -0.65,
      size = 3.1,
      show.legend = FALSE
    ) +
    scale_y_discrete(limits = rev(levels(plot_data$grade))) +
    scale_color_manual(values = c("ELA" = "#1b9e77", "Math" = "#d95f02", "Science" = "#7570b3")) +
    scale_shape_manual(
      values = c("*** p < 0.01" = 16, "** p < 0.05" = 17, "* p < 0.10" = 15, "n.s." = 1),
      drop = FALSE
    ) +
    coord_cartesian(xlim = c(-6, 14), clip = "off") +
    labs(
      title = plot_title,
      subtitle = plot_subtitle,
      x = "Adjusted JMTES coefficient in ATLAS scale-score points",
      y = NULL,
      color = "Subject",
      shape = "Appendix B significance"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      plot.margin = margin(10, 35, 10, 10)
    )
}

matching_results <- prepare_coefficient_data(
  all_results,
  "propensity_score_matching_covariate_baseline_regression"
)
did_results <- prepare_coefficient_data(
  all_results,
  "did_style_gain_covariate_baseline_regression"
)

matching_coefficient_plot <- make_coefficient_plot(
  matching_results,
  "GROW-Associated Differences in 2026 Summative ATLAS Scores: Matching Robustness Check",
  "Calipered nearest-neighbor matching with baseline and covariate adjustment; error bars show 95% CIs"
)
did_coefficient_plot <- make_coefficient_plot(
  did_results,
  "GROW-Associated Differences in 2026 Summative ATLAS Scores: Difference-in-Differences Style Check",
  "Gain-score model with baseline and covariate adjustment; error bars show 95% CIs"
)

matching_coefficient_plot
did_coefficient_plot

ggsave(
  filename = file.path(dirname(results_path), "grow_atlas_matching_coefficient_plot.png"),
  plot = matching_coefficient_plot,
  width = 10,
  height = 6,
  dpi = 300
)
ggsave(
  filename = file.path(dirname(results_path), "grow_atlas_did_coefficient_plot.png"),
  plot = did_coefficient_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```



# Balance and Common-Support Diagnostics

Because the preferred academic, attendance, and behavior designs rely on propensity-score overlap weighting, Appendix A checks whether weighting improves comparability and whether common support is plausible. The diagnostic script computes standardized mean differences before and after weighting for baseline ATLAS score (academic models), gender, race/ethnicity, age, meal-status disadvantage, entry-code indicators, and grade indicators for the attendance and behavior models. It also creates propensity-score overlap histograms by school.

```{r run-diagnostics}
# This call produces the Appendix A balance tables and propensity-score overlap figures.
system2("python3", "../run_overlap_diagnostics.py")
```

The diagnostic outputs are:

* `overlap_balance_diagnostics.csv`
* `overlap_balance_summary.csv`
* `overlap_balance_diagnostics.md`
* `propensity_score_values.csv`
* `propensity_overlap_academic.svg`
* `propensity_overlap_attendance.svg`
* `propensity_overlap_behavior.svg`
