---
title: "ATLAS GROW Impact Analysis Coefficient Plots"
author: "CSIC Evaluation Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

This R Markdown file creates coefficient plots for the ATLAS GROW impact analysis. The plots use `grow_atlas_impact_regression_results.csv` and visualize the JMTES coefficient for every available grade-by-subject model. Coefficient labels use the same significance-star convention as the report tables: * p < 0.10, ** p < 0.05, and *** p < 0.01.

The generated plot files are:

* `grow_atlas_preferred_coefficient_plot.png`
* `grow_atlas_matching_coefficient_plot.png`
* `grow_atlas_did_coefficient_plot.png`

# Shared Plot Functions

```{r coefficient-plot-functions}
library(ggplot2)

results_path <- "grow_atlas_impact_regression_results.csv"
if (!file.exists(results_path)) {
  results_path <- file.path("ATLAS GROW Impact Analysis", results_path)
}

all_results <- read.csv(results_path, stringsAsFactors = FALSE)

prepare_coefficient_data <- function(results, analysis_name) {
  plot_data <- results[results$analysis == analysis_name, ]
  plot_data$grade <- factor(
    plot_data$grade,
    levels = c("Kindergarten", "Grade 1", "Grade 2", "Grade 3", "Grade 4", "Grade 5")
  )
  plot_data$subject <- factor(plot_data$subject, levels = c("ELA", "Math", "Science"))
  plot_data$ci_lower <- plot_data$estimate - 1.96 * plot_data$std_error
  plot_data$ci_upper <- plot_data$estimate + 1.96 * plot_data$std_error
  plot_data$significance <- cut(
    plot_data$p_value,
    breaks = c(-Inf, 0.01, 0.05, 0.10, Inf),
    labels = c("*** p < 0.01", "** p < 0.05", "* p < 0.10", "n.s."),
    right = FALSE
  )
  plot_data$stars <- ifelse(
    plot_data$p_value < 0.01,
    "***",
    ifelse(plot_data$p_value < 0.05, "**", ifelse(plot_data$p_value < 0.10, "*", ""))
  )
  plot_data$estimate_label <- paste0(sprintf("%.2f", plot_data$estimate), plot_data$stars)
  plot_data
}

make_coefficient_plot <- function(plot_data, plot_title, plot_subtitle, legend_title) {
  ggplot(plot_data, aes(x = estimate, y = grade, color = subject, shape = significance)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray55", linewidth = 0.5) +
    geom_errorbarh(
      aes(xmin = ci_lower, xmax = ci_upper),
      height = 0.18,
      linewidth = 0.75,
      position = position_dodge(width = 0.55)
    ) +
    geom_point(size = 3.2, position = position_dodge(width = 0.55)) +
    geom_text(
      aes(label = estimate_label),
      position = position_dodge(width = 0.55),
      hjust = -0.65,
      size = 3.1,
      show.legend = FALSE
    ) +
    scale_y_discrete(limits = rev(levels(plot_data$grade))) +
    scale_color_manual(values = c("ELA" = "#1b9e77", "Math" = "#d95f02", "Science" = "#7570b3")) +
    scale_shape_manual(
      values = c("*** p < 0.01" = 16, "** p < 0.05" = 17, "* p < 0.10" = 15, "n.s." = 1),
      drop = FALSE
    ) +
    coord_cartesian(xlim = c(-6, 14), clip = "off") +
    labs(
      title = plot_title,
      subtitle = plot_subtitle,
      x = "Adjusted JMTES coefficient in ATLAS scale-score points",
      y = NULL,
      color = "Subject",
      shape = legend_title
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom",
      plot.margin = margin(10, 35, 10, 10)
    )
}
```

# Preferred Overlap-Weighted Model Coefficient Plot

This plot visualizes the preferred overlap-weighted, baseline-adjusted, and covariate-adjusted estimates from Table 2. It matches the main discussion by showing generally small and non-significant estimates in Kindergarten through Grade 2, consistently positive and statistically significant estimates in Grades 3 and 4, and mixed Grade 5 results with Science statistically significant.

```{r preferred-coefficient-plot, fig.width=10, fig.height=6, dpi=300}
preferred_results <- prepare_coefficient_data(
  all_results,
  "overlap_weighted_ps_covariate_baseline_regression"
)

preferred_coefficient_plot <- make_coefficient_plot(
  preferred_results,
  "GROW-Associated Differences in 2026 Summative ATLAS Scores",
  "Preferred overlap-weighted, baseline- and covariate-adjusted models; error bars show 95% CIs",
  "Table 2 significance"
)

preferred_coefficient_plot

ggsave(
  filename = file.path(dirname(results_path), "grow_atlas_preferred_coefficient_plot.png"),
  plot = preferred_coefficient_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```

# Matching Robustness Coefficient Plot

This plot visualizes the Appendix B calipered nearest-neighbor matching estimates from Table B1. It matches the Appendix B discussion by showing strongest and most consistent positive estimates in Grades 3 and 4, statistically significant Grade 5 ELA and Grade 5 Science estimates, and mostly non-significant estimates in Kindergarten through Grade 2.

```{r matching-coefficient-plot, fig.width=10, fig.height=6, dpi=300}
matching_results <- prepare_coefficient_data(
  all_results,
  "propensity_score_matching_covariate_baseline_regression"
)

matching_coefficient_plot <- make_coefficient_plot(
  matching_results,
  "GROW-Associated Differences in 2026 Summative ATLAS Scores: Matching Robustness Check",
  "Calipered nearest-neighbor matching with baseline and covariate adjustment; error bars show 95% CIs",
  "Table B1 significance"
)

matching_coefficient_plot

ggsave(
  filename = file.path(dirname(results_path), "grow_atlas_matching_coefficient_plot.png"),
  plot = matching_coefficient_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```

# Difference-in-Differences Style Coefficient Plot

This plot visualizes the Appendix B difference-in-differences style gain-score estimates from Table B2. It matches the Appendix B discussion by showing significant positive estimates concentrated in Grades 3 and 4, a significant Grade 5 Science estimate, weaker evidence in Kindergarten through Grade 2, and mixed evidence in Grade 5.

```{r did-coefficient-plot, fig.width=10, fig.height=6, dpi=300}
did_results <- prepare_coefficient_data(
  all_results,
  "did_style_gain_covariate_baseline_regression"
)

did_coefficient_plot <- make_coefficient_plot(
  did_results,
  "GROW-Associated Differences in 2026 Summative ATLAS Scores: Difference-in-Differences Style Check",
  "Gain-score model with baseline and covariate adjustment; error bars show 95% CIs",
  "Table B2 significance"
)

did_coefficient_plot

ggsave(
  filename = file.path(dirname(results_path), "grow_atlas_did_coefficient_plot.png"),
  plot = did_coefficient_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```
