---
title: "2026 Summative ATLAS Descriptive Statistics"
output:
  word_document:
    reference_docx: null
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(dplyr)
library(readr)
library(readxl)
library(knitr)
library(ggplot2)
```

```{r helper-functions}
grade_folders <- c("Kindergarten", "Grade 1", "Grade 2", "Grade 3", "Grade 4", "Grade 5")
subjects_by_grade <- list(
  "Kindergarten" = c("ELA", "Math"),
  "Grade 1" = c("ELA", "Math"),
  "Grade 2" = c("ELA", "Math"),
  "Grade 3" = c("ELA", "Math", "Science"),
  "Grade 4" = c("ELA", "Math", "Science"),
  "Grade 5" = c("ELA", "Math", "Science")
)

find_atlas_file <- function(grade, school) {
  files <- list.files(grade, pattern = paste0("^", school, ".*Atlas Report.*\\.xlsx$"), full.names = TRUE)
  current_files <- files[grepl("Current Year", files)]
  if (length(current_files) > 0) return(current_files[[1]])
  files[[1]]
}

read_atlas_file <- function(path) {
  raw <- read_excel(path, col_names = FALSE)
  header_row <- which(apply(raw, 1, function(x) any(trimws(as.character(x)) %in% c("Student ID", "Student IDs", "ID Number"))))[[1]]
  names <- as.character(unlist(raw[header_row, ]))
  dat <- raw[(header_row + 1):nrow(raw), ]
  colnames(dat) <- names
  dat
}

score_values <- function(dat, subject) {
  # Use ONLY the 2026 summative test wave for the requested subject.
  # Do not stack, average, or pool Fall/Winter/2025 scores with this column.
  score_col <- paste(subject, "Sum 2026")
  other_wave_cols <- grep(paste0("^", subject, " "), names(dat), value = TRUE)
  other_wave_cols <- setdiff(other_wave_cols, score_col)

  if (!score_col %in% names(dat)) {
    stop("Missing required 2026 summative score column: ", score_col)
  }

  scores <- suppressWarnings(as.numeric(dat[[score_col]]))
  scores[!is.na(scores)]
}

summarize_scores <- function(values) {
  # Values passed into this function are already restricted to one column:
  # the subject-specific `Sum 2026` column for one grade and school.
  values <- values[!is.na(values)]

  # Median is calculated directly from the non-missing student-level
  # 2026 summative ATLAS scores for each grade-subject-school cell.
  median_score <- median(values)

  tibble(
    n = length(values),
    mean = mean(values),
    sd = sd(values),
    median = median_score
  )
}
```

```{r calculate-table}
atlas_stats_unformatted <- bind_rows(lapply(grade_folders, function(grade) {
  jmtes <- read_atlas_file(find_atlas_file(grade, "JMTES"))
  jes <- read_atlas_file(find_atlas_file(grade, "JES"))

  bind_rows(lapply(subjects_by_grade[[grade]], function(subject) {
    jmtes_stats <- summarize_scores(score_values(jmtes, subject))
    jes_stats <- summarize_scores(score_values(jes, subject))

    tibble(
      Grade = grade,
      Subject = subject,
      jmtes_mean = jmtes_stats$mean,
      jmtes_sd = jmtes_stats$sd,
      jmtes_median = jmtes_stats$median,
      jmtes_n = jmtes_stats$n,
      jes_mean = jes_stats$mean,
      jes_sd = jes_stats$sd,
      jes_median = jes_stats$median,
      jes_n = jes_stats$n,
      difference_in_mean = jmtes_stats$mean - jes_stats$mean
    )
  }))
}))

atlas_stats <- atlas_stats_unformatted %>%
  transmute(
    Grade,
    Subject,
    `JMTES\n mean\n (SD)` = sprintf("%.2f\n (%.2f)", jmtes_mean, jmtes_sd),
    `JES\n mean\n (SD)` = sprintf("%.2f\n (%.2f)", jes_mean, jes_sd),
    `Difference-in-mean\n (JMTES - JES)` = sprintf("%.2f", difference_in_mean),
    `JMTES\n median` = format(jmtes_median, trim = TRUE, nsmall = 0),
    `JES\n median` = format(jes_median, trim = TRUE, nsmall = 0),
    `Observations\n (JMTES/JES)` = paste0(jmtes_n, "/", jes_n)
  )

write_csv(atlas_stats_unformatted, "results/atlas_2026_summative_descriptive_statistics_unformatted.csv")
write_csv(atlas_stats, "results/atlas_2026_summative_descriptive_statistics.csv")
atlas_stats
```

The `JMTES median` and `JES median` columns below are calculated with `median()` from the non-missing student-level 2026 summative ATLAS scores before the final table is formatted.

```{r display-table, results='asis'}
kable(
  atlas_stats,
  format = "pipe",
  col.names = c(
    "Grade", "Subject", "JMTES\n mean\n (SD)", "JES\n mean\n (SD)",
    "Difference-in-mean\n (JMTES - JES)", "JMTES\n median", "JES\n median",
    "Observations\n (JMTES/JES)"
  )
)
```


```{r line-plot-data}
# Build grade-level line plot data using one baseline score column and the
# subject-specific 2026 summative score column. Baseline columns are selected
# without pooling multiple waves: use the common 2025 summative baseline when
# available, otherwise use the documented same-subject baseline available in
# the file (for example, Winter 2025 for Kindergarten and Grade 3 Science).
select_baseline_col <- function(dat, grade, subject) {
  preferred_cols <- if (grade == "Kindergarten" || (grade == "Grade 3" && subject == "Science")) {
    c(paste(subject, "Winter 2025"), paste(subject, "Sum 2025"))
  } else {
    c(
      paste(subject, "Sum 2025"),
      paste(subject, "Winter 2025"),
      paste(subject, "Fall 2025"),
      paste(subject, "Sum 2024")
    )
  }

  baseline_col <- preferred_cols[preferred_cols %in% names(dat)][1]
  if (is.na(baseline_col)) {
    stop("Missing baseline score column for ", grade, " ", subject)
  }
  baseline_col
}

mean_for_col <- function(dat, col) {
  values <- suppressWarnings(as.numeric(dat[[col]]))
  mean(values, na.rm = TRUE)
}

plot_data <- bind_rows(lapply(grade_folders, function(grade) {
  jmtes <- read_atlas_file(find_atlas_file(grade, "JMTES"))
  jes <- read_atlas_file(find_atlas_file(grade, "JES"))

  bind_rows(lapply(subjects_by_grade[[grade]], function(subject) {
    outcome_col <- paste(subject, "Sum 2026")
    jmtes_baseline_col <- select_baseline_col(jmtes, grade, subject)
    jes_baseline_col <- select_baseline_col(jes, grade, subject)

    tibble(
      Grade = rep(grade, 4),
      School = rep(c("JMTES", "JES"), each = 2),
      Subject = subject,
      Wave = rep(c("Baseline", "2026 Summative"), 2),
      Wave_order = rep(c(1, 2), 2),
      Score_mean = c(
        mean_for_col(jmtes, jmtes_baseline_col),
        mean_for_col(jmtes, outcome_col),
        mean_for_col(jes, jes_baseline_col),
        mean_for_col(jes, outcome_col)
      ),
      Score_column = c(jmtes_baseline_col, outcome_col, jes_baseline_col, outcome_col)
    )
  }))
}))

write_csv(plot_data, "results/atlas_2026_summative_line_plot_data.csv")
plot_data
```

```{r line-plots, fig.width=8, fig.height=5, results='asis'}
dir.create("results/atlas_2026_summative_line_plots", showWarnings = FALSE, recursive = TRUE)

for (grade in grade_folders) {
  grade_plot_data <- plot_data %>% filter(Grade == grade)
  p <- ggplot(
    grade_plot_data,
    aes(
      x = Wave_order,
      y = Score_mean,
      color = Subject,
      shape = Subject,
      linetype = School,
      group = interaction(School, Subject)
    )
  ) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 3) +
    scale_x_continuous(
      breaks = c(1, 2),
      labels = c("Baseline", "2026 Summative")
    ) +
    scale_linetype_manual(values = c("JMTES" = "solid", "JES" = "dashed")) +
    labs(
      title = paste0(grade, ": Baseline and 2026 Summative ATLAS Mean Scores"),
      subtitle = "Color and point shape identify subject; line type identifies school",
      x = NULL,
      y = "Mean ATLAS scale score",
      color = "Subject",
      shape = "Subject",
      linetype = "School"
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")

  print(p)
  ggsave(
    filename = file.path("results/atlas_2026_summative_line_plots", paste0(gsub(" ", "_", tolower(grade)), "_atlas_line_plot.svg")),
    plot = p,
    width = 8,
    height = 5,
    units = "in"
  )
}
```

Notes: The table reports unweighted descriptive statistics for the 2026 Summative ATLAS test wave only by grade, subject, and school. Standard deviations are reported in parentheses below means. The difference-in-mean column is JMTES minus JES, so positive values indicate higher average scores at JMTES. Median values are reported separately for each school. Because the number of observations varies across grade-subject cells, observations are reported in the final column as JMTES/JES.
