---
title: "Student Incentive Distribution Analysis"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
```

## Purpose and assumptions

This R Markdown file analyzes the distribution of 2025-26 student incentives and completion of each incentive type. It assumes the student incentive tracker or merged incentive/at-risk dataset is already open in R as `student_incentives_data`.

If your open data frame has a different name, assign it before running the analysis:

```{r rename-open-dataset, eval=FALSE}
student_incentives_data <- student_incentives_at_risk_merged
```

## Configure incentive columns

```{r configure-columns}
incentive_total_col <- "2025-26 Incentive TOTAL"

incentive_columns <- tibble::tribble(
  ~incentive_type, ~aug_dec_col, ~jan_may_col,
  "Incentive 1: EA Lesson Activities",
  "Aug-Dec 2025 Incentives - Incentive 1-           EA Lesson Activities",
  "Jan-May 2026 Incentives - Incentive 1-           EA Lesson Activities",
  "Incentive 2: 1 HR Financial Literacy Module Online",
  "Aug-Dec 2025 Incentives - Incentive 2-        1 HR Fin. Literacy Module Online",
  "Jan-May 2026 Incentives - Incentive 2-        1 HR Fin. Literacy Module Online",
  "Incentive 3: Met Reading Goal",
  "Aug-Dec 2025 Incentives - Incentive 3- Met Reading Goal",
  "Jan-May 2026 Incentives - Incentive 3- Met Reading Goal",
  "Incentive 4: Met Behavior Goal",
  "Aug-Dec 2025 Incentives - Incentive 4-      Met Behavior Goal",
  "Jan-May 2026 Incentives - Incentive 4-      Met Behavior Goal",
  "Incentive 5: Miss Three Days or Less",
  "Aug-Dec 2025 Incentives - Incentive 5-        Miss three days or less",
  "Jan-May 2026 Incentives - Incentive 5-        Miss three days or less"
)

required_columns <- c(incentive_total_col, incentive_columns$aug_dec_col, incentive_columns$jan_may_col)
missing_columns <- setdiff(required_columns, names(student_incentives_data))
if (length(missing_columns) > 0) {
  stop("Missing required columns: ", paste(missing_columns, collapse = "; "))
}
```

## Total incentive dollars distributed

```{r total-incentives-distributed}
incentive_distribution_data <- student_incentives_data %>%
  mutate(incentive_total = as.numeric(.data[[incentive_total_col]]))

total_incentives_distributed <- incentive_distribution_data %>%
  summarise(
    students = n(),
    total_incentives_distributed = sum(incentive_total, na.rm = TRUE),
    average_incentive_per_student = mean(incentive_total, na.rm = TRUE)
  )

total_incentives_distributed
```

## Distribution of students by total incentive amount

```{r incentive-amount-distribution}
incentive_amount_distribution <- incentive_distribution_data %>%
  mutate(incentive_total = tidyr::replace_na(incentive_total, 0)) %>%
  count(incentive_total, name = "students") %>%
  arrange(incentive_total) %>%
  mutate(
    percent = students / sum(students),
    label = paste0("$", incentive_total, "\n", percent(percent, accuracy = 0.1))
  )

incentive_amount_distribution
```

```{r incentive-amount-pie, fig.width=7, fig.height=6}
ggplot(incentive_amount_distribution,
       aes(x = "", y = students, fill = factor(incentive_total, levels = incentive_total))) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    size = 4
  ) +
  labs(
    title = "Distribution of Students by Total 2025-26 Incentive Amount",
    fill = "Total incentive amount"
  ) +
  theme_void()
```

## Completion of each incentive type

A student is counted as completing an incentive type if they have a value of `1` in either the Aug-Dec 2025 period or the Jan-May 2026 period for that incentive.

```{r incentive-completion-analysis}
incentive_completion_long <- incentive_columns %>%
  rowwise() %>%
  do({
    tibble(
      incentive_type = .$incentive_type,
      completed = tidyr::replace_na(suppressWarnings(as.numeric(student_incentives_data[[.$aug_dec_col]])) == 1, FALSE) |
        tidyr::replace_na(suppressWarnings(as.numeric(student_incentives_data[[.$jan_may_col]])) == 1, FALSE)
    )
  }) %>%
  ungroup() %>%
  mutate(completion_status = if_else(completed, "Completed", "Not completed"))

incentive_completion_summary <- incentive_completion_long %>%
  count(incentive_type, completion_status, name = "students") %>%
  group_by(incentive_type) %>%
  mutate(
    percent = students / sum(students),
    label = paste0(students, " (", percent(percent, accuracy = 0.1), ")")
  ) %>%
  ungroup()

incentive_completion_summary
```

```{r incentive-completion-pies, fig.width=10, fig.height=8}
ggplot(incentive_completion_summary,
       aes(x = "", y = students, fill = completion_status)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    size = 3.5
  ) +
  facet_wrap(~ incentive_type) +
  labs(
    title = "Students Completing Each Incentive Type",
    fill = "Completion status"
  ) +
  theme_void() +
  theme(strip.text = element_text(face = "bold"))
```
