---
title: "Merge Student Incentive Tracker with At-Risk Report"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
library(lubridate)
```

## Purpose

This script merges the student incentive tracker to the at-risk report by Student ID. It keeps only students in the student incentive tracker and brings in all columns from the at-risk report. It also creates attendance and referral measures standardized for students who entered after the first school day.

## Assumptions

The two source datasets are already open in R as:

- `student_incentives`: the student incentive tracker.
- `at_risk_report`: the at-risk report.

If your objects have different names, rename them before running the merge code:

```{r example-object-renaming, eval=FALSE}
student_incentives <- `2025_26 Student Incentive Tracker`
at_risk_report <- `JMTES At Risk Report`
```

## Merge and derived measures

```{r merge-incentives-at-risk}
# ---- User-adjustable school year assumptions ----
# The denominator is 170 school days for students enrolled on the first school day.
# For students who entered later, enrolled days are prorated from entry date through
# the assumed last day of the measured school year. Update school_year_end if the
# at-risk report covers a different end date.
school_year_start <- as.Date("2025-08-18")
school_year_end <- as.Date("2026-05-22")
full_year_school_days <- 170

# ---- Helper functions ----
clean_student_id <- function(x) {
  x <- as.character(x)
  x <- str_trim(x)
  x <- if_else(str_detect(x, "[Ee]"),
               format(as.numeric(x), scientific = FALSE, trim = TRUE),
               x)
  str_replace_all(x, "[^0-9]", "")
}

parse_entry_date <- function(x) {
  if (inherits(x, "Date")) return(x)
  if (inherits(x, "POSIXct") || inherits(x, "POSIXt")) return(as.Date(x))

  x_chr <- str_trim(as.character(x))
  parsed <- suppressWarnings(as.Date(as.numeric(x_chr), origin = "1899-12-30"))
  parsed[is.na(parsed)] <- suppressWarnings(mdy(x_chr[is.na(parsed)]))
  parsed[is.na(parsed)] <- suppressWarnings(ymd(x_chr[is.na(parsed)]))
  parsed
}

# ---- Clean merge keys ----
student_incentives_clean <- student_incentives %>%
  mutate(student_id_clean = clean_student_id(`Student ID`))

at_risk_clean <- at_risk_report %>%
  mutate(
    student_id_clean = clean_student_id(`Student Id`),
    entry_date = parse_entry_date(`Entry Date - E/W`),
    days_enrolled_denominator = case_when(
      is.na(entry_date) ~ full_year_school_days,
      TRUE ~ pmax(
        1,
        pmin(
          full_year_school_days,
          round(((as.numeric(school_year_end - entry_date) + 1) /
                   (as.numeric(school_year_end - school_year_start) + 1)) *
                  full_year_school_days)
        )
      )
    ),
    excused_absence_percent = as.numeric(Excused) / days_enrolled_denominator,
    unexcused_absence_percent = as.numeric(Unexcused) / days_enrolled_denominator,
    total_absence_percent = (as.numeric(Excused) + as.numeric(Unexcused)) /
      days_enrolled_denominator,
    school_referrals_all_severity = rowSums(
      across(c(`L-1`, `L-2`, `L-3`, `L-4`, Local), as.numeric),
      na.rm = TRUE
    ),
    bus_referrals_all_severity = rowSums(
      across(c(`Transportation 1`, `Transportation 2`, `Transportation 3`), as.numeric),
      na.rm = TRUE
    ),
    total_referrals_all_severity = school_referrals_all_severity + bus_referrals_all_severity,
    school_referrals_per_170_days = school_referrals_all_severity /
      days_enrolled_denominator * full_year_school_days,
    bus_referrals_per_170_days = bus_referrals_all_severity /
      days_enrolled_denominator * full_year_school_days,
    total_referrals_per_170_days = total_referrals_all_severity /
      days_enrolled_denominator * full_year_school_days
  )

# ---- Merge: keep only students in the student incentive tracker ----
student_incentives_at_risk_merged <- student_incentives_clean %>%
  left_join(
    at_risk_clean,
    by = "student_id_clean",
    suffix = c("_incentive", "_at_risk")
  )

# ---- Quick merge diagnostics ----
merge_diagnostics <- student_incentives_at_risk_merged %>%
  summarise(
    incentive_tracker_students = n(),
    matched_to_at_risk_report = sum(!is.na(`Student Id`)),
    unmatched_to_at_risk_report = sum(is.na(`Student Id`)),
    match_rate = matched_to_at_risk_report / incentive_tracker_students
  )

merge_diagnostics
```

## Export merged dataset

```{r export-merged-data, eval=FALSE}
write.csv(
  student_incentives_at_risk_merged,
  "Incentive Tracker/student_incentive_tracker_merged_at_risk.csv",
  row.names = FALSE
)
```
