---
title: "Financial Literacy Mean Difference and 95% Confidence Interval Plots"
author: "GROW Evaluation"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
  word_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(scipen = 999, digits = 3)
```

## Purpose

This R Markdown file uses `all_grades_financial_literacy_scores.csv` to visualize paired pre-post financial literacy score changes. The figure reports mean paired differences and 95% confidence intervals for two outcomes:

- **Raw-score change**, shown separately for Grade 3, Grade 4, and Grade 5.
- **Percent-correct change**, shown separately for Grade 3, Grade 4, Grade 5, and all students combined.

Positive mean differences indicate higher post-test scores than pre-test scores. The confidence intervals are calculated from student-level paired change scores using the standard error of the mean difference and the appropriate two-sided *t* critical value.

## Load packages and data

```{r load-data}
library(ggplot2)

scores <- read.csv("all_grades_financial_literacy_scores.csv", stringsAsFactors = FALSE)

scores$grade_label <- factor(
  paste("Grade", scores$grade),
  levels = c("Grade 3", "Grade 4", "Grade 5")
)
```

## Create paired difference measures

The raw-score paired difference is calculated as post-test total correct minus pre-test total correct. Percent-correct scores place grade-level assessments on a common scale. Grade 3 had 10 scored items, while Grades 4 and 5 had 11 scored items; therefore, percent-correct scores are calculated with grade-specific denominators.

```{r create-differences}
scores$max_score <- ifelse(scores$grade == 3, 10, 11)

scores$pre_percent_correct <- 100 * scores$pre_total_correct / scores$max_score
scores$post_percent_correct <- 100 * scores$post_total_correct / scores$max_score

scores$raw_score_difference <- scores$post_total_correct - scores$pre_total_correct
scores$percent_correct_difference <- scores$post_percent_correct - scores$pre_percent_correct
```

## Helper function for mean differences and 95% confidence intervals

```{r helper-functions}
mean_ci <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  mean_difference <- mean(x)
  sd_difference <- sd(x)
  se_difference <- sd_difference / sqrt(n)
  t_critical <- qt(0.975, df = n - 1)

  data.frame(
    n = n,
    mean_difference = mean_difference,
    se_difference = se_difference,
    ci_low = mean_difference - t_critical * se_difference,
    ci_high = mean_difference + t_critical * se_difference
  )
}
```

## Summarize paired changes

```{r summarize-changes}
raw_by_grade <- do.call(
  rbind,
  lapply(split(scores, scores$grade_label), function(dat) {
    cbind(
      sample = as.character(dat$grade_label[1]),
      outcome = "Raw score",
      mean_ci(dat$raw_score_difference)
    )
  })
)

percent_by_grade <- do.call(
  rbind,
  lapply(split(scores, scores$grade_label), function(dat) {
    cbind(
      sample = as.character(dat$grade_label[1]),
      outcome = "Percent correct",
      mean_ci(dat$percent_correct_difference)
    )
  })
)

percent_all <- cbind(
  sample = "All grades",
  outcome = "Percent correct",
  mean_ci(scores$percent_correct_difference)
)

plot_data <- rbind(raw_by_grade, percent_by_grade, percent_all)

plot_data$outcome <- factor(plot_data$outcome, levels = c("Raw score", "Percent correct"))
plot_data$sample <- factor(
  plot_data$sample,
  levels = c("Grade 3", "Grade 4", "Grade 5", "All grades")
)

plot_data
```

## Paired-samples t-test table

```{r paired-ttest-table}
paired_ttest_summary <- function(dat, difference_variable) {
  difference_scores <- dat[[difference_variable]]
  difference_scores <- difference_scores[!is.na(difference_scores)]
  test_result <- t.test(difference_scores, mu = 0)

  data.frame(
    n = length(difference_scores),
    mean_difference = mean(difference_scores),
    se_difference = sd(difference_scores) / sqrt(length(difference_scores)),
    t_statistic = unname(test_result$statistic),
    p_value = test_result$p.value,
    ci_low = test_result$conf.int[1],
    ci_high = test_result$conf.int[2]
  )
}

raw_ttests <- do.call(
  rbind,
  lapply(split(scores, scores$grade_label), function(dat) {
    cbind(
      panel = "Panel A. Raw score paired t-test results",
      sample = as.character(dat$grade_label[1]),
      paired_ttest_summary(dat, "raw_score_difference")
    )
  })
)

percent_ttests_by_grade <- do.call(
  rbind,
  lapply(split(scores, scores$grade_label), function(dat) {
    cbind(
      panel = "Panel B. Percent-correct paired t-test results",
      sample = as.character(dat$grade_label[1]),
      paired_ttest_summary(dat, "percent_correct_difference")
    )
  })
)

percent_ttests_all <- cbind(
  panel = "Panel B. Percent-correct paired t-test results",
  sample = "All grades",
  paired_ttest_summary(scores, "percent_correct_difference")
)

paired_ttests <- rbind(raw_ttests, percent_ttests_by_grade, percent_ttests_all)
paired_ttests
```

## Visualization: mean paired differences with 95% confidence intervals

```{r mean-difference-ci-plot, fig.width=10, fig.height=6}
ggplot(plot_data, aes(x = sample, y = mean_difference)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.15, linewidth = 0.8) +
  geom_point(size = 3, color = "#1f78b4") +
  geom_text(
    aes(label = sprintf("%.2f", mean_difference)),
    nudge_y = 0.6,
    size = 3.5,
    color = "#1f78b4"
  ) +
  facet_wrap(~ outcome, scales = "free_y") +
  labs(
    title = "Mean Pre-Post Financial Literacy Score Differences",
    subtitle = "Points show mean paired differences; bars show 95% confidence intervals",
    x = NULL,
    y = "Mean difference (post-test minus pre-test)",
    caption = "Raw-score results are shown within grade only because grade-level assessments used different denominators. Percent-correct results include within-grade estimates and a combined all-students estimate."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 30, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

## Interpretation

The plotted mean differences match the paired-samples framework: each student's post-test score is compared with that same student's pre-test score, and the average student-level change is shown with a 95% confidence interval. Confidence intervals that cross zero indicate that the observed average gains are not statistically distinguishable from no mean change at the conventional 5% level. Therefore, the figure should be interpreted as descriptive evidence of positive score movement rather than statistically significant evidence of improvement.
