---
title: "Comparing Behavioral Referral Outcomes Between JMTES and JES"
author: "GROW Evaluation"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(width = 120, scipen = 999)
```

# Purpose

This analysis estimates the association between attendance at Jacksonville Middle/Elementary School (JMTES, the GROW school) and behavioral referral outcomes using Jacksonville Elementary School (JES) as the comparison school. The analysis follows the overlap-weighted propensity score strategy used in the attendance analysis, but it uses Poisson regression because referral outcomes are nonnegative counts with many zeroes.

All Poisson models include an exposure offset for enrollment days so that students are compared on referral rates rather than raw counts alone. The coefficient on `jmtes` is reported as an incidence rate ratio (IRR). IRRs below 1 indicate lower referral rates at JMTES relative to comparable JES students, while IRRs above 1 indicate higher referral rates at JMTES.

# Data sources

```{r data-sources}
jmtes_file <- file.path("..", "JMTES At Risk Report with Absence Percentages.csv")
jes_file <- file.path("..", "JES At Risk Report with Absence Percentages.csv")

required_files <- c(jmtes_file, jes_file)
missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files) > 0) {
  stop("Missing required file(s): ", paste(missing_files, collapse = ", "))
}
```

# Helper functions

```{r helper-functions}
clean_chr <- function(x) trimws(as.character(x))
num <- function(x) suppressWarnings(as.numeric(clean_chr(x)))

format_p <- function(p) {
  ifelse(is.na(p), NA_character_, ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))
}

stars <- function(p) {
  ifelse(p < 0.01, "***", ifelse(p < 0.05, "**", ifelse(p < 0.10, "*", "")))
}

robust_poisson_row <- function(model, data, outcome, label) {
  x <- model.matrix(model)
  w <- model.weights(model.frame(model))
  mu <- fitted(model)
  y <- model.response(model.frame(model))
  response_resid <- y - mu

  bread <- solve(t(x) %*% (x * as.numeric(w * mu)))
  meat <- t(x) %*% (x * as.numeric((w^2) * (response_resid^2)))
  robust_vcov <- bread %*% meat %*% bread

  estimate <- unname(coef(model)["jmtes"])
  robust_se <- sqrt(unname(robust_vcov["jmtes", "jmtes"]))
  z_value <- estimate / robust_se
  p_value <- 2 * pnorm(abs(z_value), lower.tail = FALSE)
  irr <- exp(estimate)
  ci_low <- exp(estimate - 1.96 * robust_se)
  ci_high <- exp(estimate + 1.96 * robust_se)

  weighted_rate <- function(group) {
    idx <- data$jmtes == group
    sum(data$overlap_weight[idx] * data[[outcome]][idx]) /
      sum(data$overlap_weight[idx] * data$enrollment_days[idx]) * 170
  }

  data.frame(
    outcome = outcome,
    label = label,
    coefficient = estimate,
    coefficient_ci_low = estimate - 1.96 * robust_se,
    coefficient_ci_high = estimate + 1.96 * robust_se,
    robust_se = robust_se,
    z_value = z_value,
    p_value = p_value,
    irr = irr,
    ci_low = ci_low,
    ci_high = ci_high,
    percent_difference = 100 * (irr - 1),
    percent_ci_low = 100 * (ci_low - 1),
    percent_ci_high = 100 * (ci_high - 1),
    jes_weighted_rate_per_170_days = weighted_rate(0),
    jmtes_weighted_rate_per_170_days = weighted_rate(1),
    observations = nrow(data),
    events = sum(data[[outcome]]),
    sparse_or_unstable = sum(data[[outcome]]) < 10 || weighted_rate(0) == 0 || weighted_rate(1) == 0,
    row.names = NULL
  )
}

robust_logit_row <- function(model, data, outcome, label) {
  x <- model.matrix(model)
  w <- model.weights(model.frame(model))
  mu <- fitted(model)
  y <- model.response(model.frame(model))
  response_resid <- y - mu

  bread <- solve(t(x) %*% (x * as.numeric(w * mu * (1 - mu))))
  meat <- t(x) %*% (x * as.numeric((w^2) * (response_resid^2)))
  robust_vcov <- bread %*% meat %*% bread

  estimate <- unname(coef(model)["jmtes"])
  robust_se <- sqrt(unname(robust_vcov["jmtes", "jmtes"]))
  z_value <- estimate / robust_se
  p_value <- 2 * pnorm(abs(z_value), lower.tail = FALSE)
  odds_ratio <- exp(estimate)
  ci_low <- exp(estimate - 1.96 * robust_se)
  ci_high <- exp(estimate + 1.96 * robust_se)

  weighted_probability <- function(group) {
    idx <- data$jmtes == group
    sum(data$overlap_weight[idx] * data[[outcome]][idx]) / sum(data$overlap_weight[idx])
  }

  data.frame(
    outcome = outcome,
    label = label,
    coefficient = estimate,
    robust_se = robust_se,
    z_value = z_value,
    p_value = p_value,
    odds_ratio = odds_ratio,
    ci_low = ci_low,
    ci_high = ci_high,
    jes_weighted_probability = weighted_probability(0),
    jmtes_weighted_probability = weighted_probability(1),
    observations = nrow(data),
    events = sum(data[[outcome]]),
    row.names = NULL
  )
}
```

# Import and prepare referral outcomes

```{r import-clean}
read_school <- function(file, school_name, jmtes_indicator) {
  raw <- read.csv(file, stringsAsFactors = FALSE, check.names = FALSE)

  l1 <- num(raw[["L-1"]])
  l2 <- num(raw[["L-2"]])
  l3 <- num(raw[["L-3"]])
  l4 <- num(raw[["L-4"]])
  local <- num(raw[["Local"]])
  bus1 <- num(raw[["Transportation 1"]])
  bus2 <- num(raw[["Transportation 2"]])
  bus3 <- num(raw[["Transportation 3"]])

  data.frame(
    school = school_name,
    jmtes = jmtes_indicator,
    student_id = clean_chr(raw[["Student Id"]]),
    grade = clean_chr(raw[["Student Grade"]]),
    female = ifelse(tolower(clean_chr(raw[["Gender"]])) == "female", 1, 0),
    black = ifelse(grepl("^black$", tolower(clean_chr(raw[["Ethnic Name"]]))), 1, 0),
    age = num(raw[["Student Age"]]),
    meal_status = clean_chr(raw[["Meal Status Code"]]),
    entry_code = clean_chr(raw[["Entry Code - E/W"]]),
    enrollment_days = num(raw[["Report School Days"]]),
    classroom_total = l1 + l2 + l3 + l4,
    classroom_l1 = l1,
    classroom_l2 = l2,
    classroom_l3 = l3,
    classroom_l4 = l4,
    classroom_severity_points = l1 + 2 * l2 + 3 * l3 + 4 * l4,
    bus_total = local + bus1 + bus2 + bus3,
    bus_l1 = local + bus1,
    bus_l2 = bus2,
    bus_l3 = bus3,
    bus_severity_points = local + bus1 + 2 * bus2 + 3 * bus3,
    stringsAsFactors = FALSE
  )
}

referrals <- rbind(
  read_school(jmtes_file, "JMTES", 1),
  read_school(jes_file, "JES", 0)
)

referrals <- subset(
  referrals,
  !is.na(jmtes) & !is.na(grade) & !is.na(female) & !is.na(black) & !is.na(age) &
    !is.na(meal_status) & !is.na(entry_code) & !is.na(enrollment_days) & enrollment_days > 0
)

referrals$school <- factor(referrals$school, levels = c("JES", "JMTES"))
referrals$grade <- factor(referrals$grade)
referrals$female <- as.numeric(referrals$female)
referrals$black <- as.numeric(referrals$black)
referrals$meal_status <- factor(referrals$meal_status)
referrals$entry_code <- factor(referrals$entry_code)

aggregate(enrollment_days ~ school, data = referrals, FUN = function(x) unique(x)[1])
```

# Propensity score and overlap weights

```{r propensity-model}
ps_model <- glm(
  jmtes ~ grade + female + black + age + meal_status + entry_code,
  data = referrals,
  family = binomial()
)

referrals$propensity <- as.numeric(predict(ps_model, type = "response"))
referrals$overlap_weight <- ifelse(referrals$jmtes == 1, 1 - referrals$propensity, referrals$propensity)

summary(referrals$propensity)
summary(referrals$overlap_weight)
```

# Overlap-weighted Poisson referral models

Each model uses the same covariate-adjusted Poisson specification with robust standard errors and an offset for enrollment days:

\[
log(E[Referrals_i]) = \beta_0 + \beta_1JMTES_i + \beta_2Grade_i + \beta_3Female_i + \beta_4Black_i + \beta_5Age_i + \beta_6MealStatus_i + \beta_7EntryCode_i + log(EnrollmentDays_i)
\]

```{r poisson-models}
outcomes <- c(
  classroom_total = "Classroom total referrals",
  classroom_l1 = "Classroom Level I referrals",
  classroom_l2 = "Classroom Level II referrals",
  classroom_l3 = "Classroom Level III referrals",
  classroom_l4 = "Classroom Level IV referrals",
  classroom_severity_points = "Classroom severity-weighted frequency",
  bus_total = "Bus total referrals",
  bus_l1 = "Bus/Local Level I referrals",
  bus_l2 = "Bus Level II referrals",
  bus_l3 = "Bus Level III referrals",
  bus_severity_points = "Bus severity-weighted frequency"
)

fit_referral_model <- function(outcome) {
  model <- glm(
    as.formula(paste(outcome, "~ jmtes + grade + female + black + age + meal_status + entry_code")),
    data = referrals,
    family = poisson(link = "log"),
    weights = overlap_weight,
    offset = log(enrollment_days)
  )
  robust_poisson_row(model, referrals, outcome, unname(outcomes[outcome]))
}

referral_results <- do.call(rbind, lapply(names(outcomes), fit_referral_model))
referral_results$p_value_formatted <- format_p(referral_results$p_value)
referral_results$stars <- stars(referral_results$p_value)
referral_results
```

# Robustness check: any-referral logistic models

Because the referral outcomes have medians of zero, differences in mean referral counts could be influenced by a smaller number of students with multiple referrals. This robustness check estimates overlap-weighted logistic regression models for whether a student had at least one referral. These models test whether JMTES is associated with the extensive-margin probability of receiving any referral, rather than only the average number of referrals.

```{r any-referral-logit-robustness}
referrals$any_school_referral <- as.numeric(referrals$classroom_total > 0)
referrals$any_bus_referral <- as.numeric(referrals$bus_total > 0)
referrals$any_referral <- as.numeric(referrals$classroom_total > 0 | referrals$bus_total > 0)

any_referral_outcomes <- c(
  any_school_referral = "Any school referral",
  any_bus_referral = "Any bus referral",
  any_referral = "Any school or bus referral"
)

fit_any_referral_model <- function(outcome) {
  model <- glm(
    as.formula(paste(outcome, "~ jmtes + grade + female + black + age + meal_status + entry_code")),
    data = referrals,
    family = binomial(link = "logit"),
    weights = overlap_weight
  )
  robust_logit_row(model, referrals, outcome, unname(any_referral_outcomes[outcome]))
}

any_referral_results <- do.call(rbind, lapply(names(any_referral_outcomes), fit_any_referral_model))
any_referral_results$p_value_formatted <- format_p(any_referral_results$p_value)
any_referral_results$stars <- stars(any_referral_results$p_value)
any_referral_results
```

# AER-style regression table

```{r aer-table}
make_aer_table <- function(results, outcome_names) {
  selected <- results[match(names(outcome_names), results$outcome), ]
  irr_text <- ifelse(selected$sparse_or_unstable, "Sparse/unstable", paste0(sprintf("%.3f", selected$irr), selected$stars))
  se_text <- ifelse(selected$sparse_or_unstable, "--", paste0("(", sprintf("%.3f", selected$robust_se), ")"))
  ci_text <- ifelse(
    selected$sparse_or_unstable,
    "--",
    paste0("[", sprintf("%.3f", selected$ci_low), ", ", sprintf("%.3f", selected$ci_high), "]")
  )
  percent_text <- ifelse(selected$sparse_or_unstable, "--", paste0(sprintf("%.1f", selected$percent_difference), "%"))
  table <- rbind(
    `JMTES IRR` = irr_text,
    `Robust SE` = se_text,
    `95% CI` = ci_text,
    `Percent difference` = percent_text,
    `JES weighted rate per 170 days` = sprintf("%.2f", selected$jes_weighted_rate_per_170_days),
    `JMTES weighted rate per 170 days` = sprintf("%.2f", selected$jmtes_weighted_rate_per_170_days),
    Observations = as.character(selected$observations)
  )
  colnames(table) <- unname(outcome_names)
  table
}

classroom_table <- make_aer_table(referral_results, outcomes[grep("^classroom", names(outcomes))])
bus_table <- make_aer_table(referral_results, outcomes[grep("^bus", names(outcomes))])

classroom_table
bus_table
```

# Coefficient plot

```{r coefficient-plot, fig.width=10, fig.height=7}
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  stop("The ggplot2 package is required to create the coefficient plot.")
}

plot_data <- referral_results
plot_data$domain <- ifelse(grepl("^classroom", plot_data$outcome), "Classroom", "Bus")
plot_data$label <- factor(plot_data$label, levels = rev(plot_data$label))

referral_coefficient_plot <- ggplot2::ggplot(
  plot_data,
  ggplot2::aes(x = coefficient, y = label, xmin = coefficient_ci_low, xmax = coefficient_ci_high)
) +
  ggplot2::geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  ggplot2::geom_errorbarh(height = 0.18, linewidth = 0.8, color = "#2C7FB8") +
  ggplot2::geom_point(size = 3, color = "#2C7FB8") +
  ggplot2::facet_wrap(~ domain, scales = "free_y") +
  ggplot2::labs(
    title = "Overlap-weighted Poisson estimates for referral outcomes",
    subtitle = "Points are log incidence rate ratios for JMTES relative to JES; bars are 95% confidence intervals",
    x = "JMTES coefficient (log incidence rate ratio)",
    y = NULL
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(panel.grid.minor = ggplot2::element_blank())

referral_coefficient_plot
```

# Export outputs

```{r export}
write.csv(referrals, "jmtes_jes_referral_student_data.csv", row.names = FALSE)
write.csv(referral_results, "jmtes_jes_referral_poisson_results.csv", row.names = FALSE)
write.csv(any_referral_results, "jmtes_jes_referral_any_referral_logit_robustness_results.csv", row.names = FALSE)
write.csv(as.data.frame(classroom_table), "jmtes_jes_referral_aer_classroom_table.csv")
write.csv(as.data.frame(bus_table), "jmtes_jes_referral_aer_bus_table.csv")

ggplot2::ggsave(
  filename = "jmtes_jes_referral_coefficient_plot.svg",
  plot = referral_coefficient_plot,
  width = 10,
  height = 7
)
```

# Discussion guide

The main estimates are the JMTES IRRs in the AER-style tables. For total classroom referrals and total bus referrals, IRRs below 1 indicate that comparable JMTES students had lower referral rates than comparable JES students. The severity-weighted outcomes evaluate whether the overall frequency of more serious referrals differs after assigning larger weights to higher-severity incidents. These results remain quasi-experimental because the design adjusts for observed student characteristics but does not randomly assign students to schools.
