---
title: "Referral Overlap Histograms"
author: "GROW Evaluation"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(width = 120, scipen = 999)
```

# Purpose

This report creates overlapped histograms for referral-count distributions at JMTES and JES. The y-axis is the percentage of students within each school who fall in each referral-count bin, so the two school distributions can be compared even though the schools have different sample sizes.

The school referral figure includes total school referrals and Levels I-IV. The bus referral figure includes total bus referrals and Levels I-III.

# Packages and data

```{r packages-data}
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  stop("The ggplot2 package is required to render these histograms.")
}

library(ggplot2)

student_data_file <- "jmtes_jes_referral_student_data.csv"
if (!file.exists(student_data_file)) {
  stop("Missing ", student_data_file, ". Run run_referral_analysis.py first, then render this R Markdown file from the Referral Evaluation folder.")
}

referral_data <- read.csv(student_data_file, stringsAsFactors = FALSE, check.names = FALSE)
referral_data$school <- factor(referral_data$school, levels = c("JMTES", "JES"))
```

# Helper functions

```{r helpers}
make_histogram_data <- function(data, outcomes) {
  pieces <- lapply(seq_len(nrow(outcomes)), function(i) {
    variable <- outcomes$variable[i]
    label <- outcomes$label[i]
    max_count <- max(data[[variable]], na.rm = TRUE)
    bins <- 0:max_count

    do.call(rbind, lapply(levels(data$school), function(school_name) {
      values <- data[[variable]][data$school == school_name]
      values <- values[!is.na(values)]
      counts <- tabulate(match(values, bins), nbins = length(bins))
      data.frame(
        outcome = label,
        count_bin = bins,
        students = counts,
        percent_students = 100 * counts / length(values),
        school = school_name,
        stringsAsFactors = FALSE
      )
    }))
  })

  histogram_data <- do.call(rbind, pieces)
  histogram_data$outcome <- factor(histogram_data$outcome, levels = outcomes$label)
  histogram_data$school <- factor(histogram_data$school, levels = c("JMTES", "JES"))
  histogram_data
}

plot_overlap_histograms <- function(histogram_data, title) {
  ggplot(
    histogram_data,
    aes(x = count_bin, y = percent_students, fill = school, color = school)
  ) +
    geom_col(position = "identity", alpha = 0.45, width = 0.85) +
    facet_wrap(~ outcome, scales = "free_x") +
    scale_x_continuous(breaks = function(x) seq(floor(x[1]), ceiling(x[2]), by = 1)) +
    scale_y_continuous(labels = function(x) paste0(x, "%"), expand = expansion(mult = c(0, 0.08))) +
    scale_fill_manual(values = c("JMTES" = "#1f77b4", "JES" = "#ff7f0e")) +
    scale_color_manual(values = c("JMTES" = "#1f77b4", "JES" = "#ff7f0e")) +
    labs(
      title = title,
      x = "Referral count",
      y = "Students in bin",
      fill = "School",
      color = "School"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      strip.text = element_text(face = "bold"),
      legend.position = "top",
      panel.grid.minor = element_blank()
    )
}
```

# School referral histograms

```{r school-histograms, fig.width=12, fig.height=8}
school_outcomes <- data.frame(
  variable = c("classroom_total", "classroom_l1", "classroom_l2", "classroom_l3", "classroom_l4"),
  label = c(
    "Total school referrals",
    "School Level I referrals",
    "School Level II referrals",
    "School Level III referrals",
    "School Level IV referrals"
  ),
  stringsAsFactors = FALSE
)

school_histogram_data <- make_histogram_data(referral_data, school_outcomes)
school_histogram_plot <- plot_overlap_histograms(
  school_histogram_data,
  "School Referral Distributions by School"
)
school_histogram_plot

ggsave(
  "jmtes_jes_school_referral_overlap_histograms_ggplot.svg",
  school_histogram_plot,
  width = 12,
  height = 8,
  units = "in"
)
```

# Bus referral histograms

```{r bus-histograms, fig.width=12, fig.height=7}
bus_outcomes <- data.frame(
  variable = c("bus_total", "bus_l1", "bus_l2", "bus_l3"),
  label = c(
    "Total bus referrals",
    "Bus Level I referrals",
    "Bus Level II referrals",
    "Bus Level III referrals"
  ),
  stringsAsFactors = FALSE
)

bus_histogram_data <- make_histogram_data(referral_data, bus_outcomes)
bus_histogram_plot <- plot_overlap_histograms(
  bus_histogram_data,
  "Bus Referral Distributions by School"
)
bus_histogram_plot

ggsave(
  "jmtes_jes_bus_referral_overlap_histograms_ggplot.svg",
  bus_histogram_plot,
  width = 12,
  height = 7,
  units = "in"
)
```
