---
title: "Comparing School Absences Between JMTES and JES"
author: "GROW Evaluation"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
  word_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(width = 120, scipen = 999)
```

# Purpose

This analysis estimates the association between attendance at Jacksonville Middle/Elementary School (JMTES, the GROW school) and student absence outcomes, using Jacksonville Elementary School (JES) as the comparison school. Because students were not randomly assigned to JMTES or JES, the primary specification uses an overlap-weighted propensity score design. Overlap weighting emphasizes students with similar observed baseline characteristics across schools and down-weights students whose covariate profiles are uncommon in the other school.

The outcomes are absence percentages rather than raw absence counts. This is important because the JMTES file was collected before the school year ended, when the report covered 159 school days, whereas the JES file covered 170 school days. Percentage outcomes therefore place both schools on a comparable scale.

# Data sources

```{r data-sources}
jmtes_file <- "JMTES At Risk Report with Absence Percentages.csv"
jes_file   <- "JES At Risk Report with Absence Percentages.csv"

required_files <- c(jmtes_file, jes_file)
missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files) > 0) {
  stop("Missing required file(s): ", paste(missing_files, collapse = ", "))
}
```

# Helper functions

```{r helper-functions}
clean_chr <- function(x) trimws(as.character(x))

num <- function(x) suppressWarnings(as.numeric(clean_chr(x)))

weighted_mean <- function(x, w) {
  ok <- is.finite(x) & is.finite(w) & w > 0
  sum(w[ok] * x[ok]) / sum(w[ok])
}

weighted_var <- function(x, w) {
  ok <- is.finite(x) & is.finite(w) & w > 0
  x <- x[ok]
  w <- w[ok]
  mu <- sum(w * x) / sum(w)
  sum(w * (x - mu)^2) / sum(w)
}

smd_numeric <- function(x, treat, w = NULL) {
  if (is.null(w)) w <- rep(1, length(x))
  m1 <- weighted_mean(x[treat == 1], w[treat == 1])
  m0 <- weighted_mean(x[treat == 0], w[treat == 0])
  v1 <- weighted_var(x[treat == 1], w[treat == 1])
  v0 <- weighted_var(x[treat == 0], w[treat == 0])
  denom <- sqrt((v1 + v0) / 2)
  if (!is.finite(denom) || denom == 0) return(NA_real_)
  (m1 - m0) / denom
}

smd_binary <- function(x, treat, w = NULL) {
  if (is.null(w)) w <- rep(1, length(x))
  p1 <- weighted_mean(as.numeric(x)[treat == 1], w[treat == 1])
  p0 <- weighted_mean(as.numeric(x)[treat == 0], w[treat == 0])
  p <- (p1 + p0) / 2
  denom <- sqrt(p * (1 - p))
  if (!is.finite(denom) || denom == 0) return(NA_real_)
  (p1 - p0) / denom
}

make_factor_smds <- function(data, var, treat, w = NULL) {
  lvls <- sort(unique(as.character(data[[var]])))
  do.call(rbind, lapply(lvls, function(level) {
    indicator <- as.character(data[[var]]) == level
    data.frame(
      variable = paste0(var, ": ", level),
      smd = smd_binary(indicator, treat, w),
      row.names = NULL
    )
  }))
}

format_p <- function(p) {
  ifelse(is.na(p), NA_character_, ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))
}

model_row <- function(model, term = "jmtes") {
  coefs <- coef(summary(model))
  out <- coefs[term, , drop = FALSE]
  data.frame(
    estimate = unname(out[1, "Estimate"]),
    std_error = unname(out[1, "Std. Error"]),
    t_value = unname(out[1, "t value"]),
    p_value = unname(out[1, "Pr(>|t|)"]),
    conf_low = unname(confint(model, term)[1]),
    conf_high = unname(confint(model, term)[2]),
    row.names = NULL
  )
}
```

# Import and prepare data

```{r import-clean}
read_school <- function(file, school_name, jmtes_indicator) {
  raw <- read.csv(file, stringsAsFactors = FALSE, check.names = FALSE)
  data.frame(
    school = school_name,
    jmtes = jmtes_indicator,
    student_id = clean_chr(raw[["Student Id"]]),
    grade = clean_chr(raw[["Student Grade"]]),
    female = ifelse(tolower(clean_chr(raw[["Gender"]])) == "female", 1, 0),
    gender = clean_chr(raw[["Gender"]]),
    black = ifelse(grepl("^black$", tolower(clean_chr(raw[["Ethnic Name"]]))), 1, 0),
    race_ethnicity = clean_chr(raw[["Ethnic Name"]]),
    age = num(raw[["Student Age"]]),
    meal_status = clean_chr(raw[["Meal Status Code"]]),
    entry_code = clean_chr(raw[["Entry Code - E/W"]]),
    report_school_days = num(raw[["Report School Days"]]),
    full_school_year_days = num(raw[["Full School Year Days"]]),
    unexcused_absences = num(raw[["Unexcused"]]),
    excused_absences = num(raw[["Excused"]]),
    total_absences = num(raw[["Total Absences"]]),
    unexcused_pct = num(raw[["Unexcused Absence %"]]),
    excused_pct = num(raw[["Excused Absence %"]]),
    total_pct = num(raw[["Total Absence %"]]),
    stringsAsFactors = FALSE
  )
}

attendance <- rbind(
  read_school(jmtes_file, "JMTES", 1),
  read_school(jes_file, "JES", 0)
)

attendance <- subset(
  attendance,
  !is.na(jmtes) & !is.na(grade) & !is.na(female) & !is.na(black) & !is.na(age) &
    !is.na(meal_status) & !is.na(entry_code) & !is.na(unexcused_pct) & !is.na(excused_pct) & !is.na(total_pct)
)

attendance$school <- factor(attendance$school, levels = c("JES", "JMTES"))
attendance$grade <- factor(attendance$grade)
attendance$gender <- factor(attendance$gender)
attendance$female <- as.numeric(attendance$female)
attendance$black <- as.numeric(attendance$black)
attendance$meal_status <- factor(attendance$meal_status)
attendance$entry_code <- factor(attendance$entry_code)

school_summary <- aggregate(
  cbind(student_id = attendance$student_id, report_school_days = attendance$report_school_days),
  by = list(school = attendance$school),
  FUN = function(x) if (is.character(x)) length(unique(x)) else unique(x)[1]
)
names(school_summary) <- c("school", "students", "report_school_days")
school_summary
```

# Descriptive absence percentages

```{r descriptive-outcomes}
outcome_summary <- do.call(rbind, lapply(c("unexcused_pct", "excused_pct", "total_pct"), function(v) {
  do.call(rbind, lapply(levels(attendance$school), function(s) {
    x <- attendance[attendance$school == s, v]
    data.frame(
      outcome = v,
      school = s,
      n = length(x),
      mean = mean(x),
      sd = sd(x),
      median = median(x),
      q25 = unname(quantile(x, 0.25)),
      q75 = unname(quantile(x, 0.75)),
      row.names = NULL
    )
  }))
}))
outcome_summary
```

# Propensity score model

The propensity score model estimates each student's probability of attending JMTES rather than JES based on observed covariates available in the at-risk report files:

\[
p_i = Pr(JMTES_i = 1 \mid X_i)
\]

\[
logit(p_i) = \alpha_0 + \alpha_1 Grade_i + \alpha_2 Female_i + \alpha_3 Black_i + \alpha_4 Age_i + \alpha_5 MealStatus_i + \alpha_6 EntryCode_i
\]

```{r propensity-model}
ps_model <- glm(
  jmtes ~ grade + female + black + age + meal_status + entry_code,
  data = attendance,
  family = binomial()
)

attendance$propensity <- as.numeric(predict(ps_model, type = "response"))
attendance$overlap_weight <- ifelse(attendance$jmtes == 1, 1 - attendance$propensity, attendance$propensity)

summary(attendance$propensity)
summary(attendance$overlap_weight)
```

# Covariate balance

Standardized mean differences (SMDs) are shown before and after overlap weighting. Smaller absolute SMDs indicate better balance between JMTES and JES. As a rule of thumb, absolute SMDs below 0.10 are often considered well balanced.

```{r balance}
treat <- attendance$jmtes
balance_rows <- rbind(
  data.frame(variable = "age", unweighted_smd = smd_numeric(attendance$age, treat), weighted_smd = smd_numeric(attendance$age, treat, attendance$overlap_weight)),
  data.frame(variable = "female", unweighted_smd = smd_binary(attendance$female == 1, treat), weighted_smd = smd_binary(attendance$female == 1, treat, attendance$overlap_weight)),
  data.frame(variable = "black", unweighted_smd = smd_binary(attendance$black == 1, treat), weighted_smd = smd_binary(attendance$black == 1, treat, attendance$overlap_weight))
)

factor_vars <- c("grade", "meal_status", "entry_code")
factor_balance <- do.call(rbind, lapply(factor_vars, function(v) {
  unweighted <- make_factor_smds(attendance, v, treat)
  weighted <- make_factor_smds(attendance, v, treat, attendance$overlap_weight)
  data.frame(variable = unweighted$variable, unweighted_smd = unweighted$smd, weighted_smd = weighted$smd)
}))

balance_table <- rbind(balance_rows, factor_balance)
balance_table$abs_unweighted_smd <- abs(balance_table$unweighted_smd)
balance_table$abs_weighted_smd <- abs(balance_table$weighted_smd)
balance_table <- balance_table[order(-balance_table$abs_unweighted_smd), ]
balance_table
```

# Overlap-weighted outcome models

For each absence percentage outcome, the primary model is:

\[
Y_i = \beta_0 + \beta_1 JMTES_i + \epsilon_i
\]

The doubly adjusted model adds the same covariates used in the propensity score model:

\[
Y_i = \beta_0 + \beta_1 JMTES_i + \beta_2 Grade_i + \beta_3 Female_i + \beta_4 Black_i + \beta_5 Age_i + \beta_6 MealStatus_i + \beta_7 EntryCode_i + \epsilon_i
\]

In both cases, \(\beta_1\) is the estimated difference in absence percentage points for JMTES students relative to comparable JES students. Negative estimates indicate lower absence percentages at JMTES; positive estimates indicate higher absence percentages at JMTES.

```{r weighted-models}
outcomes <- c(
  unexcused_pct = "Unexcused absence percentage",
  excused_pct = "Excused absence percentage",
  total_pct = "Total absence percentage"
)

fit_outcome <- function(outcome) {
  primary_formula <- as.formula(paste(outcome, "~ jmtes"))
  adjusted_formula <- as.formula(paste(outcome, "~ jmtes + grade + female + black + age + meal_status + entry_code"))

  primary_model <- lm(primary_formula, data = attendance, weights = overlap_weight)
  adjusted_model <- lm(adjusted_formula, data = attendance, weights = overlap_weight)

  primary <- model_row(primary_model)
  adjusted <- model_row(adjusted_model)

  primary$model <- "Overlap weighted"
  adjusted$model <- "Overlap weighted + covariates"
  out <- rbind(primary, adjusted)
  out$outcome <- unname(outcomes[outcome])
  out
}

model_results <- do.call(rbind, lapply(names(outcomes), fit_outcome))
model_results <- model_results[, c("outcome", "model", "estimate", "std_error", "t_value", "p_value", "conf_low", "conf_high")]
model_results$p_value_formatted <- format_p(model_results$p_value)
model_results
```

The regression table above is the main reporting table. It includes all three requested outcomes and both model specifications: the overlap-weighted model without covariates and the overlap-weighted model with the same covariates used in the propensity score model.

The JMTES coefficients from the model without covariates and the model with covariates are expected to be very similar when the propensity score and overlap weights are working correctly. For overlap weights estimated from a logistic propensity score, the weighted covariate means for the variables in the propensity score model are balanced across the treatment and comparison groups. Because the covariates are already balanced after weighting, adding those same covariates to the weighted outcome regression should not substantially change the JMTES coefficient. Similar standard errors are also expected when the added covariates explain little additional weighted residual variation beyond the school indicator and the overlap-weighted design.

```{r model-comparison-check}
model_comparison_check <- reshape(
  model_results[, c("outcome", "model", "estimate", "std_error")],
  idvar = "outcome",
  timevar = "model",
  direction = "wide"
)
names(model_comparison_check) <- gsub("estimate\\.", "estimate_", names(model_comparison_check))
names(model_comparison_check) <- gsub("std_error\\.", "std_error_", names(model_comparison_check))

model_comparison_check$estimate_difference <- with(
  model_comparison_check,
  `estimate_Overlap weighted + covariates` - `estimate_Overlap weighted`
)
model_comparison_check$std_error_difference <- with(
  model_comparison_check,
  `std_error_Overlap weighted + covariates` - `std_error_Overlap weighted`
)
model_comparison_check
```

# Weighted means by school

```{r weighted-means}
weighted_means <- do.call(rbind, lapply(names(outcomes), function(v) {
  do.call(rbind, lapply(levels(attendance$school), function(s) {
    idx <- attendance$school == s
    data.frame(
      outcome = unname(outcomes[v]),
      school = s,
      weighted_mean = weighted_mean(attendance[idx, v], attendance$overlap_weight[idx]),
      unweighted_mean = mean(attendance[idx, v]),
      row.names = NULL
    )
  }))
}))
weighted_means
```

# Export analysis tables

```{r export}
write.csv(attendance, "jmtes_jes_overlap_weighted_attendance_student_data.csv", row.names = FALSE)
write.csv(balance_table, "jmtes_jes_overlap_weighted_attendance_balance.csv", row.names = FALSE)
write.csv(model_results, "jmtes_jes_overlap_weighted_attendance_results.csv", row.names = FALSE)
write.csv(weighted_means, "jmtes_jes_overlap_weighted_attendance_weighted_means.csv", row.names = FALSE)
write.csv(model_comparison_check, "jmtes_jes_overlap_weighted_attendance_model_comparison_check.csv", row.names = FALSE)

regression_table <- model_results
regression_table$estimate_label <- sprintf(
  "%.3f (SE = %.3f), 95%% CI [%.3f, %.3f], p = %s",
  regression_table$estimate,
  regression_table$std_error,
  regression_table$conf_low,
  regression_table$conf_high,
  regression_table$p_value_formatted
)
write.csv(regression_table, "jmtes_jes_overlap_weighted_attendance_regression_table.csv", row.names = FALSE)

xml_escape <- function(x) {
  x <- as.character(x)
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  x <- gsub(">", "&gt;", x, fixed = TRUE)
  x <- gsub("\"", "&quot;", x, fixed = TRUE)
  x
}

xml_row <- function(values) {
  cells <- vapply(values, function(value) {
    if (is.numeric(value) && length(value) == 1 && is.finite(value)) {
      paste0("<Cell><Data ss:Type=\"Number\">", value, "</Data></Cell>")
    } else {
      paste0("<Cell><Data ss:Type=\"String\">", xml_escape(value), "</Data></Cell>")
    }
  }, character(1))
  paste0("<Row>", paste(cells, collapse = ""), "</Row>")
}

regression_table_for_excel <- regression_table
regression_table_for_excel$p_value <- regression_table_for_excel$p_value_formatted

excel_xml <- c(
  "<?xml version=\"1.0\"?>",
  "<?mso-application progid=\"Excel.Sheet\"?>",
  "<Workbook xmlns=\"urn:schemas-microsoft-com:office:spreadsheet\"",
  " xmlns:o=\"urn:schemas-microsoft-com:office:office\"",
  " xmlns:x=\"urn:schemas-microsoft-com:office:excel\"",
  " xmlns:ss=\"urn:schemas-microsoft-com:office:spreadsheet\"",
  " xmlns:html=\"http://www.w3.org/TR/REC-html40\">",
  "<Worksheet ss:Name=\"Regression Table\"><Table>",
  xml_row(names(regression_table_for_excel)),
  apply(regression_table_for_excel, 1, xml_row),
  "</Table></Worksheet>",
  "<Worksheet ss:Name=\"Notes\"><Table>",
  xml_row("Notes"),
  xml_row("The JMTES coefficient is measured in absence percentage points relative to JES."),
  xml_row("No-covariate models regress each absence percentage outcome on the JMTES indicator using overlap weights."),
  xml_row("Covariate-adjusted models add grade, female, Black, age, meal-status code, and entry code."),
  xml_row("Overlap weights are 1 - propensity score for JMTES students and propensity score for JES students."),
  xml_row("Percentage outcomes are used because JMTES was reported through 159 school days and JES through 170 school days."),
  xml_row("The no-covariate and covariate-adjusted JMTES coefficients are expected to be similar when overlap weighting balances the same covariates that are later added to the outcome model."),
  "</Table></Worksheet>",
  "</Workbook>"
)
writeLines(excel_xml, "jmtes_jes_overlap_weighted_attendance_regression_table_excel.xml")
```

# Interpretation guide

The key estimates are the JMTES coefficients in `model_results`. Each estimate is measured in absence percentage points. For example, an estimate of `-1.25` for total absence percentage would mean that, after overlap weighting, JMTES students missed 1.25 percentage points fewer enrolled/report days than comparable JES students. Because this is a quasi-experimental comparison rather than a randomized design, the estimates should be interpreted as adjusted associations that depend on the observed covariates included in the propensity score and outcome models.
